---
title: 
csl: plos.csl
fontsize: 10pt
output:
  pdf_document:
    includes:
      in_header: latex/header.tex
    keep_tex: yes
  html_document:
    df_print: paged
link-citations: yes
geometry: left=2cm, right=2cm, top=2cm, bottom=3cm, footskip = .5cm
subparagraph: yes
bibliography: SOE2020.bib
---

```{r setup, include=FALSE}

# library(tint)
# # invalidate cache when the package version changes
# knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
# options(htmltools.dir.version = FALSE)

#Default Rmd options
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      dev = "cairo_pdf",
                      warning = FALSE,
                      #fig.width = 4,
                      #fig.asp = 0.45,
                      fig.align = 'center') #allows for inserting R code into captions

#Plotting and data libraries
#remotes::install_github("noaa-edab/ecodata@0.1.0") #change to 2020 ecodata version for release
library(tidyverse)
library(tidyr)
library(ecodata)
library(here)
library(kableExtra)

```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-setup.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-GIS-setup.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-setup.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-setup.R")}
```

\setcounter{page}{3}

# Characterizing ecosystem change for fishery management

The Northeast shelf ecosystem is changing and affecting the services the ecosystem provides. To illustrate how multiple factors are driving change in this complex ecosystem we are using three overarching concepts: multiple system drivers, regime shifts, and ecosystem reorganization. Physical, chemical, biological, and human factors comprise the **multiple system drivers** that influence the quality and distribution of habitat in the ocean, which affects the amount and diversity of fish in the ocean. These environmental drivers combined with multiple social and economic drivers influence the range of fishing opportunities and the seafood, recreation, and other ecosystem services we derive from the ocean. 

```{r, out.width='72%'}

knitr::include_url("https://github.com/NOAA-EDAB/presentations/raw/master/docs/EDAB_images/multipledrivers.png")

```

Changes in those drivers can lead to **regime shifts** — large, abrupt and persistent changes in the structure and function of an ecosystem. Regime shifts and changes in how the multiple system drivers interact can result in **ecosystem reorganization** as species and humans respond and adapt to the new environment. 

```{r, out.width='72%'}

knitr::include_url("https://github.com/NOAA-EDAB/presentations/raw/master/docs/EDAB_images/RegimeReorg.png")
```

# Report Structure

The purpose of this report is to synthesize ecosystem information to better meet fishery management objectives, and to update the MAFMC’s Ecosystem Approach to Fishery Management (EAFM) risk assessment. The major messages of the report are synthesized in the 2-page summary. The information in this report is organized into two sections; performance against ecosystem-level management objectives (Table \ref{tab:management-objectives}), and risks to fishery management (Climate change and Other Ocean Uses).  

Detailed technical methods documentation^[https://NOAA-EDAB.github.io/tech-doc] and indicator data^[https://github.com/NOAA-EDAB/ecodata] are available online. The details of standard figure formatting (Fig. \ref{fig:docformat}a), categorization of fish and invertebrate species into feeding groups (Table \ref{tab:species-groupings}), and definitions of ecological production units (EPUs, including the Mid-Atlantic Bight, MAB; Fig. \ref{fig:docformat}b) are provided at the end of the document. 

```{r management-objectives}

mng_obj <- data.frame("Objective Categories" = c("Seafood Production",
                                                 "Profits","Recreation",
                                                 "Stability","Social & Cultural",
                                                 "Protected Species",
                                                 "Biomass","Productivity",
                                                 "Trophic structure","Habitat"),
"Indicators reported here" = c("Landings by feeding guild","Revenue decomposed to price and volume",
                               "Days fished; recreational catch",
                               "Diversity indices (fishery and species)",
                               "Community engagement/reliance status",
                               "",
                               "Biomass or abundance by feeding guild from surveys",
                               "Condition and recruitment of managed species, Primary productivity",
                               "Relative biomass of feeding guilds, Zooplankton",
                               "Estuarine and offshore habitat conditions"))

knitr::kable(mng_obj, linesep = "",
      col.names = c("Objective Categories","Indicators reported here"),
      caption = "Established ecosystem-scale objectives in the Mid-Atlantic Bight",
      #align = 'c',
      booktabs = T) %>%
  kable_styling(latex_options = "hold_position", "scale_down") %>%
 # column_spec(c(2), width = c("25em")) %>%
  row_spec(0, bold = TRUE) %>%
 # group_rows("Provisioning/Cultural", 1,4) %>%
 # group_rows("Supporting/Regulating", 5,9)
  pack_rows("Provisioning/Cultural Services", 1,6) %>%
  pack_rows("Supporting/Regulating Services", 7,10)


```
 
# Performance relative to fishery management objectives

In this section, we examine indicators related to broad, ecosystem-level fishery management objectives. We also provide hypotheses about *why* we are seeing these trends by identifying interacting drivers (other indicators in this report), and where possible, any regime shifts or potential changes in ecosystem structure. Identifying multiple drivers, regime shifts, and potential changes to ecosystem structure, as well as mapping most vulnerable resources, can help managers determine whether we can do anything differently to meet objectives and how to prioritize for upcoming issues/risks. 

## Seafood Production 

### Indicator: Landings

Total commercial landings and MAFMC's managed landings in the MAB continue to trend down. 

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-comdat-total-landings.R"), fig.height=1.5}
```

The downward trend is most significant in the benthos (clams and scallops) group.

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-comdat-commercial-landings.R"), fig.height=5}
```

Recreational landings are also down overall in the MAB.
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-recdat-landings.R"), fig.height=1.5}

```

Recreational shark landings show increased catch of pelagic sharks over the past decade, with a sharp decrease in the two most recent years.
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-rec_hms.R"), fig.height=1.5}

```

Aquaculture production is not yet included in total seafood landings, but we are working towards that in future reports. Available aquaculture production of oysters for a subset of Mid-Atlantic states is trending upwards^[https://noaa-edab.github.io/catalog/aquaculture.html]

### Why

Not ecosystem overfishing
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-fogarty.R"), fig.height=2}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-ryther.R"), fig.height=2}
```

Not stock status: most ok
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-stock-status.R"), fig.height=4}
```

Not biomass: most trends are steady, spring piscivores and fall benthos show long term increase (although many surveys not completed in 2020, NEAMAP Fall is included)
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-aggregate-biomass.R"), fig.height=5}
```

Commercial landings decline:  
- Benthos landings decline likely driven by market dynamics  
    - Surf clam and ocean quahog not binding quotas  
    - Downward price pressure from international supply [reported in performance reports](https://static1.squarespace.com/static/511cdc7fe4b00307a2628ac6/t/5f075b6737cbe041406404ab/1594317671964/f_FPR_for2020_SurfclamOceanQuahog.pdf), increasing costs of aging fleet/closures  
    - Climate change increases mixing of 2 species, regulatory restrictions on catching a mix  
    - Shifting into deeper water and northward.  
    
Recreational landings decline: 
- Recreational shark landings declining due to management (previously less regulated)
- Most recent lows in all other recreational fish could still be the change in MRIP survey, calibration not quite lining up. Limits actually increased here so less likely management  
- Won't know till we get more data points  
- 80s/90s vs recent rec: could also be survey instrument? Could also be change from less management to more active management  
- Decline in rec effort since 2010, maybe generational shift, not at same rate for catch/landings  

Even if market driven, changes in other indicators require monitoring as they may become important drivers of landings in the future:  
- Climate trending to uncharted territory (see [Climate Risks section](#climate-and-ecosystem-productivity))  
- Stocks are shifting distribution throughout the region  
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-species-dist.R"), fig.height=3}
```  
- Some ecosystem composition and production changes (see [Stability section](#stability))  
- Engagement declining (see [Social and Cultural section](#social-and-cultural))  

## Profits 

### Indicators: revenue (a proxy for profits), with price and volume components  
Total commercial revenue has increased over the long term but the trend may be reversing,with recent total revenue below the long term average. MAFMC managed species revenue continued its downward trend, with recent years near a time series low.
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-comdat-comm-revenue.R"), fig.height=1.5}
```

Revenue trends overall mirror those for the benthos (clams and scallops) group over the past decade.  
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-bennet2.R"), fig.height=3}
```

### Why  
Increasing revenue is due to increasing quantities landed, which offset declining prices. Recent declines in prices contributed to falling revenue as quantities landed did not increase enough to overcome declining prices. Falling prices for surf clams and ocean quahogs in particular were likely due to increased competition from international sources, according to the [fishery performance report](https://static1.squarespace.com/static/511cdc7fe4b00307a2628ac6/t/5f075b6737cbe041406404ab/1594317671964/f_FPR_for2020_SurfclamOceanQuahog.pdf).

Even if market driven, changes in other indicators require monitoring as they may become important drivers of revenue in the future:   
-  Surf clams and ocean quahogs are vulnerable to climate change because of their sensitivity to warming ocean temperatures and ocean acidification.   
-  New observations of acidification in surf clam summer habitat are approaching, but not yet at, levels affecting surf clam growth (see [Climate Risks section](#climate-and-ecosystem-productivity)).  

## Recreational Opportunities 

### Indicators: Days fished, fleet diversity  
Recreational days fished have been stable over the long term, with current days fished near the long term average. However, recreational fleet diversity has declined over the long term.
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-recdat-effort.R"), fig.height=1.5}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-recdat-diversity.R"), fig.height=1.5}

```
- Overall effort being steady but change in diversity of recreational fleet effort suggests a changing range of opportunities
- Changes access to types of species that can be caught
- impacts?

### Why
Multiple drivers
-  Shift from charter party to shore based over the long term
    -  Shoreside larger component
    -  Changing species caught
(ASMFS and SAFMC managed species maintain catch diversity--next section, will break up plots)


## Stability 

### Indicators: fishery fleet and catch diversity, ecological component diversity
Commercial fishery fleet count, fleet diversity and permit diversity has been stable over time in the MAB with current values near the long term average. This indicates similar commercial fleet composition and species targeting opportunities over time. 
-  Reliant on a smaller number of species relative to mid 90s, but now consistent and near long term average
-  Good to increase diversity or bad? Losing traditional fisheries, cascading effects on newly entered fisheries?

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-commercial-div.R"), fig.height=4}
```

As noted [above](#recreational-opportunities), recreational fleet effort diversity is unstable (declining). However, recreational species catch diversity is stable.
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-recdat-div-catch.R"), fig.height=1.5}
```

Zooplankton diversity is increasing in the MAB, while larval fish and adult fish diversity indices are stable over time with current values near the long term average.

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-zoo-diversity.R"), fig.height=1.5}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-ichthyo-diversity.R"), fig.height=1.5}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-exp-n.R"), fig.height=2}
```

 
### Why
Stability in commercial fleets reflects...  adequate capacity to respond to current fishery changes? no big shocks from management or the ecosystem so far? permitting structure that enforces similar composition? 

Declining recreational fleet effort diversity, as noted [above](#recreational-opportunities), indicates that the contraction by the party/charter boat sector continues, with shoreside angling becoming more important, by percentage.

Stability in recreational species catch diversity has been maintained by recent increase in ASMFC and SAFMC managed species.

Increasing zooplankton diversity is driven by ... more southern species (tropicalization?) or the loss of a dominant species (Cal fin?) with everything else staying about the same?

Stable larval and adult fish diversity suggests ... the same overall number and evenness over time, but doesn't rule out species substitutions (e.g., warm-water replacing cold-water). Stable but variable (larval) can indicate changes in a dominant species. 


## Social and Cultural 

### Indicators: Commercial and recreational engagement and reliance by community
In 2020, we reported that the number of highly engaged Mid-Atlantic commercial fishing communities have declined over time, and the engagement scores have also declined in medium-highly engaged communities. Here we focus on engagement and reliance scores for individual communities, as well as social vulnerability. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-commercial-engagement.R"), fig.height=4}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-recreational-engagement.R"), fig.height=4}
```


### Why
List multiple drivers of community status: landings, permits, etc (definition of engagement and reliance), as well as drivers of social vulnerability.

Provide statement of how managers can use this status plot going forward

## Protected Species
Protected species include marine mammals (under the Marine Mammal Protection Act), endangered and threatened species (under the Endangered Species Act), and migratory birds (under the Migratory Bird Treaty Act). In the Northeast US, endangered/threatened species include Atlantic salmon, Atlantic and shortnose sturgeon, all sea turtle species, and 5 baleen whales. Fishery management objectives for protected species generally focus on reducing threats and on habitat conservation/restoration; here we report on the status of these actions as well as indicating the potential for future interactions driven by observed and predicted ecosystem changes in the Northeast US region. 

Protected species objectives include managing bycatch to remain below potential biological removal (PBR) thresholds, recovering endangered populations, and monitoring unusual mortality events (UMEs). 

### Indicators: bycatch, population (adult and juvenile) numbers, mortalities
Both harbor porpoise and gray seal bycatch indices are below current PBR thresholds, meeting management objectives. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-harborporpoise.R"), fig.height=2}
```

(Blue line here is specific to New England--did we get Mid specific estimates?)
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-grayseal.R"), fig.height=2}
```

The North Atlantic right whale population was on a recovery trajectory until 2010, but has since declined. Reduced survival rates of adult females and diverging abundance trends between sexes have also been observed. It is estimated that there are only about 100 adult females remaining in the population. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-narw-abundance.R"), fig.height=2}
```
North Atlantic right whale calf counts have also been declining. In 2018 there were zero observed new calves, and a drop in annual calves roughly mirrors the abundance decline (Fig above), however seven new calves were born in 2019. Preliminary data suggest that 2020 calf count is not 0; 12 calves have been observed as of January 2021.
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-NARW-calf-abundance.R"), fig.height=2}
```

This year saw the continuation of four Unusual Mortality Events (UMEs). Three large whale Unusual Mortality Events (UMEs) are ongoing for North Atlantic right whales, humpback whales, and minke whales. In all three cases human interaction appears to have contributed to increased mortalities, although investigations are not complete. Since 2017, the total UME right whale mortalities includes 30 dead stranded whales, 9 in the US and 21 in Canada. During 2019 9 whales have been stranded in Canada and one in the US, 3 of these were determined to have due to vessel strike while the remainder are undetermined at this time. A UME for both gray and harbor seals was declared in 2018, triggering an investigation into the cause of this event. Tests so far suggest phocine distemper virus as a potential cause, although the investigation is not yet complete. 


### Why
Bycatch management measures have been implemented. The downward trend in harbor porpoise bycatch can also be due to changing seasonal distributions of harbor porpoise and fisheries reducing overlap and a decrease in gill net effort. The increasing trend in gray seal bycatch is related to an increase in the gray seal population (US pup counts). 

The number of grey seals (Halichoerus grypus) in U.S. waters has risen dramatically in the last 2 decades, with few observed in the early 1990s to roughly 24,000 observed in southeastern Massachusetts in 2015. Roughly 30,000 - 40,000 gray seals were estimated in southeastern Massachusetts in 2015, using correction factors applied to seal counts visible in Google Earth imagery. As of 2016, the size of the grey seal population in Canada, which is part of the same stock as the grey seals in the U.S., was estimated to be roughly 425,000, and increasing by 4% a year. In U.S. waters, the number of pupping sites has increased from 1 in 1988 to 9 in 2019. Mean rates of increase in the number of pups born at various times since 1988 at 4 of the more data-rich pupping sites (Muskeget, Monomoy, Seal, and Green Islands) ranged from -0.2% (95%CI: -2.3 - 1.9%) to 26.3% (95%CI: 21.6 - 31.4%). These high rates of increase provide further support that seals from Canada are continually supplementing the breeding population in U.S. waters.

Strong evidence exists to suggest that the level of interaction between right whales and the combination of offshore lobster gear in the US and snow crab gear in Canada is contributing substantially to the decline of the species. Further, right whale distribution has changed since 2010. The reasons for these changes is unclear, but changes in climate and primary prey (Calanus finmarchicus) are suspected. 

A marine mammal climate vulnerability assessment is currently underway and for Atlantic and Gulf of Mexico populations and will be reported on in future versions of this report. 

# Risks to meeting fishery management objectives

## Climate and Ecosystem Productivity

### Climate Change Indicators: 


### Ecosystem Productivity Indicators:


### Implications


## Other Ocean Uses: Offshore Wind

### Indicators: revenue in lease areas, development timeline, survey overlap
Commercial fishery revenue in proposed offshore wind lease areas ranges from x-y% of total revenue for MAFMC managed species.
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-wind-revenue.R"), fig.height=2}
```

Offshore wind projects may cover more than 1.5 million acres by 2029.
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-wind-dev-speed0.R"), fig.height=2}
```

Scientific survey area overlap with offshore wind lease areas ranges from 2-25% across ecosystem, shellfish, fish, and protected species surveys.
```{r}
knitr::include_url("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/docs/images/SurveysWindFarmAreas_20201229.jpg")
```

### Implications
Fishery revenue could be displaced to other areas or foregone if all wind projects are built.  
Displaced fishing effort can alter fishing methods and change habitat, species (managed and protected), and fleet interactions.  
Right whales may be displaced and could face possible prey impacts via changed oceanography.  
Current plans for rapid buildout of offshore wind in a patchwork of areas spreads the impacts differentially throughout the region.  
```{r}
knitr::include_url("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/docs/images/MidAtlantic_2021119-01.jpg")
```

Offshore wind will pose a risk to scientific data collection for surveys and increased uncertainty for management decision making across physical, low trophic, fish, and protected species.




# Contributors

**Editors** (NOAA NMFS Northeast Fisheries Science Center, NEFSC): Sarah Gaichas, Kimberly Bastille, Geret DePiper, Kimberly Hyde, Scott Large, Sean Lucey, Chris Orphanides 

**Contributors** (NEFSC unless otherwise noted): Andy Beet, Patricia Clay, Lisa Colburn, Geret DePiper, Michael Fogarty, Paula Fratantoni, Kevin Friedland, Sarah Gaichas, Avijit Gangopadhyay (School for Marine Science and Technology, University of Massachusetts Dartmouth), James Gartland (Virginia Institute of Marine Science), Glen Gawarkiewicz (Woods Hole Oceanographic Institution), Sean Hardison, Kimberly Hyde, John Kosik, Steve Kress and Don Lyons (National Audubon Society’s Seabird Restoration Program), Ruth Boettcher (Virginia Department of Game and Inland Fisheries), Young-Oh Kwon and Zhuomin Chen (Woods Hole Oceanographic Institute), Sean Lucey, Chris Melrose, Ryan Morse, Kimberly Murray, Chris Orphanides, Richard Pace, Charles Perretti, Vincent Saba, Laurel Smith, John Walden, Harvey Walsh, Mark Wuenschel.  


\newpage 

# Document Orientation

The figure format is illustrated in Fig \ref{fig:docformat}a. Trend lines are shown when slope is significantly different from 0 at the p < 0.05 level. An orange line signifies an overall positive trend, and purple signifies a negative trend. To minimize bias introduced by small sample size, no trend is fit for < 30 year time series. Dashed lines represent mean values of time series unless the indicator is an anomaly, in which case the dashed line is equal to 0. Shaded regions indicate the past ten years. If there are no new data for 2018, the shaded region will still cover this time period. The spatial scale of indicators is either coastwide, Mid-Atlantic states (New York, New Jersey, Delaware, Maryland, Virginia, North Carolina), or at the Mid-Atlantic Bight (MAB) Ecosystem Production Unit (EPU, Fig. \ref{fig:docformat}b) level.

```{r docformat, fig.cap = "Document orientation. a. Key to figures. b.The Northeast Large Marine Ecosystem.",  fig.width = 8, fig.height = 2.5}
#fig.subcap= c('Key to figures.', 'The Northeast Large Marine Ecosystem.'), out.width = '.49\\linewidth', fig.show="hold" 
# FIgure orientation subfigure
m <- 0.1
x <- 1985:2018
y <-  m*x + rnorm(30, sd = 0.35)

data <- data.frame(x = x,
                  y = y)

#Define constants for figure plot
x.shade.max <- max(x)
x.shade.min <- x.shade.max - 9 
hline = mean(y)

#Plot series with trend 
psample <- ggplot2::ggplot(data = data,aes(x = x, y = y)) +
  #Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  geom_point(size = pcex) +
  scale_color_manual(aesthetics = "color")+
  guides(color = FALSE) +
  geom_hline(aes(yintercept = hline),
              size = hline.size,
             alpha = hline.alpha,
             linetype = hline.lty)+
  geom_line() +
  geom_gls() +
  scale_y_continuous(labels = function(l){trans = l / 1000})+
  scale_x_continuous(breaks = seq(1985, 2015, by = 5), expand = c(0.01, 0.01)) +
  ylab(expression("Invented Index, 10"^3*"widgets")) +
  xlab("Time") +
  labs(tag = "a")  +
  theme_ts()


# EPU map subfigure
# Set lat/lon window for map
xmin = -78
xmax = -65
ymin = 36
ymax = 45
xlims <- c(xmin, xmax)
ylims <- c(ymin, ymax)

#CRS
crs <- "+proj=longlat +lat_1=35 +lat_2=45 +lat_0=40 +lon_0=-77 +x_0=0 +y_0=0 +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"

#Specify data frame with lat/lon locations for labels
epu_labels <- data.frame(EPU = c("Mid-Atlantic\n Bight",
                                 "Gulf of Maine",
                                 "Georges Bank"),
                         latitude = c(40,42.85,41),
                         longitude = c(-72.7,-69,-68.5))

#Map of NE LME
epumap <- ggplot() +
  geom_sf(data = coast, size = map.lwd) +
  geom_sf(data = epu_sf, fill = "transparent", size = map.lwd) +
  coord_sf(crs = crs, xlim = xlims, ylim = ylims) +
  geom_text(data = epu_labels, aes(x = longitude,
                                    y = latitude,
                                    label = EPU),
            size = 1.7)+
  theme_map() +
  scale_x_continuous(breaks = seq(-78, -65, by = 4), expand = c(0.01, 0.01)) +
  xlab("Longitude") +
  ylab("Latitude") +
  labs(tag = "b")

psample + plot_spacer() + epumap + plot_layout(ncol = 3, widths=c(3,.5,4.5))

```

Fish and invertebrates are aggregated into similar feeding categories (Table \ref{tab:species-groupings}) to evaluate ecosystem level trends in predators and prey. 

```{r species-groupings, warning=F}

# new table with all species listed by management entity
df <- ecodata::species_groupings %>%
  dplyr::select(SOE.20, COMNAME, Fed.Managed) %>%
  filter(SOE.20 != "Other") %>%
  distinct() %>%
  group_by(SOE.20, Fed.Managed) %>%
  summarize_all(funs(paste(na.omit(.), collapse = ", "))) %>%
  spread(Fed.Managed, COMNAME) %>%
  arrange(factor(SOE.20, levels = c("Apex Predator", "Piscivore", "Planktivore", "Benthivore", "Benthos")))
df<-df[c(1,3,2,4,5)] %>%
  mutate_all(tolower)


knitr::kable(df, booktabs = TRUE, caption = 'Feeding guilds and management bodies.', 
             col.names = c("Guild", "MAFMC", "Joint", "NEFMC", "State or Other")) %>%
  kable_styling(font_size=10, latex_options=c("repeat_header", "scale_down", "hold_position")) %>%
  row_spec(0,bold=TRUE) %>%
  column_spec(1, width="2cm") %>%
  column_spec(2, width="4cm") %>%
  column_spec(3, width="2cm") %>%
  column_spec(4, width="5cm") %>%
  column_spec(5, width="6cm") %>%
  #column_spec(3, width="7.5cm") #%>%
  collapse_rows(columns = 1, latex_hline = "major", valign = "middle")

```


\newpage 

# References
