---
title: "State of the Ecosystem 2019: Mid-Atlantic"
output: pdf_document
---

```{r setup, include=FALSE}

#Default Rmd options
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = 'center') #allows for inserting R code into captions

#Plotting and data libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(ecodata)
library(here)
library(kableExtra)
library(ggrepel)
library(stringr)
library(patchwork)
library(grid)
library(ggiraph)
library(vegan)
library(rpart)

#GIS libraries
library(sf)
library(rgdal)
library(raster)
library(rnaturalearth)

#Data directories
image.dir <- here("images")
gis.dir <- here("gis")

#GIS directory
#gis.dir <- here::here("inst","extdata","gridded")

#General inline text input for report
#Council
council <- "Mid-Atlantic Fishery Management Council"
council_abbr <- "MAFMC"

#Region identifiers
epu <- "Mid-Atlantic Bight"
epu_abbr <- "MAB"
region <- "Mid-Atlantic"
region_abbr <- "MA" #Some commercial data organized by "MA" or "NE" regions, not by EPU 


```


```{r GIS-setup}

#CRS
crs <- "+proj=longlat +lat_1=35 +lat_2=45 +lat_0=40 +lon_0=-77 +x_0=0 +y_0=0 +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"

#Coastline shapefile
coast <- ne_countries(scale = 10,
                          continent = "North America",
                          returnclass = "sf") %>%
             sf::st_transform(crs = crs)

#State polygons
ne_states <- ne_states(country = "united states of america",
                                      returnclass = "sf") %>%
  sf::st_transform(crs = crs)

#high-res polygon of Maine
#new_england <- read_sf(gis.dir,"new_england")

#EPU shapefile
epu_sf <- ecodata::epu_sf %>% 
  filter(EPU %in% c("MAB","GB","GOM"))

#Map line parameters
map.lwd <- 0.4

# Set lat/lon window for maps
xmin = -77
xmax = -65
ymin = 36
ymax = 45
xlims <- c(xmin, xmax)
ylims <- c(ymin, ymax)

#Time series constants
shade.alpha <- 0.3
shade.fill <- "lightgrey"
lwd <- 1
pcex <- 2
trend.alpha <- 0.5
trend.size <- 2
hline.size <- 1
hline.alpha <- 0.35
hline.lty <- "dashed"
label.size <- 5
hjust.label <- 1.5
letter_size <- 4
feeding.guilds <- c("Apex Predator","Piscivore","Planktivore","Benthivore","Benthos")
x.shade.min <- 2009
x.shade.max <- 2018
#Function for custom ggplot facet labels
label <- function(variable,value){
  return(facet_names[value])
}
```

# Take Home Messages

The purpose of this report is to synthesize available information relevant to fishery management in the Mid-Atlantic portion of the US Northeast Shelf. This 2019 report highlights where management interventions have proven successful to achieve ecological objectives, but also the considerable challenges for management posed by climate change and increasing trade-offs across conservation, fishing, and other human activities in this region. Finally, we describe combinations of ecological signals that present opportunities for further integrated research and possibly creative management solutions. 

Some management interventions for ecosystems and fisheries have proven beneficial in the Mid-Atlantic and throughout the Northeast US Shelf. For example, management limiting nutrient inputs has significantly improved water quality in Chesapeake Bay (fig.X). This trend, which has been driven by improvements to dissolved oxygen content and water clarity, has positive implications for aquaculture development and for managed species using the Bay as nursery habitat.  Also encouraging is a continued decrease in harbor porpoise bycatch, which for this year is the lowest on record in the region, and an apparent increase in the harbor porpoise population (p. X).

However, there are multiple signals indicating challenges to meeting management goals. Despite mostly meeting fishery management objectives at the single species level (fig.X), long term declines in total seafood production and commercial revenue remain apparent (figX). Indicators highlight a declining diversity of recreational opportunities (fishing modes and species, figX). Further, coastal communities with high fishery engagement and reliance are dependent on a smaller number of species than historically (pX), these species are predominantly high valued shellfish vulnerable to increased ocean temperature and acidification. 

Fisheries are significantly affecting protected species, particularly large whale abundance. Changes in fishery management to address this issue could have considerable consequences for the fixed gear fishing sector, particularly for pot gear such as for lobster. The recovery of north Atlantic right whales is jeopardized by anthropogenic mortality (figX). Unusual Mortality Events (UMEs) have been declared for north Atlantic right whales, humpback whales, and minke whales, and preliminary investigations suggest fishing gear entanglement has played a primary role in these mortalities (pX). Gray seal bycatch mortality has also risen in recent years, with an estimated annual mortality that has often totaled more than 1000 animals and a UME has been declared for both gray and harbor seals that may be due to phocine distemper virus (pX). Shifts in ecosystem conditions may be contributing to increased interaction between right whales and fishing gear and could be playing a role in other protected species trends.

Several climate and ecosystem observations are trending towards unprecedented levels.  Sea surface and bottom temperatures across the Northeast US shelf are still among the fastest warming waters globally (figX) with implications for species physiology, productivity, distribution, and community composition. Globally, 2018 was the 4th hottest year on record and the last four years were the warmest on record. In the Mid-Atlantic, 2018 summer sea surface temperatures were the third highest on record, however temperatures during the other seasons were near or slightly below average.  Bottom temperature measurements also continue to have a significant long term increasing trend (figX). Ocean circulation is changing as well. The position of the northern edge of the Gulf Stream has been moving to the north since the late 1950’s, with an increasing rate since 2009. The most northerly positions ever recorded were observed over the most recent years (2014-2017, figX). A more northerly Gulf Stream position is associated with warmer ocean temperature in the Northeast US shelf (Zhang and Vallis, 2007) and increased sea surface height along the U.S. east coast (Goddard et al. 2015). 

The management implications of these ocean changes vary by region and species, and are not fully understood at this point. Changes in the distribution of managed fish species continue, with aggregate trends on the entire Northeast Shelf towards the northeast and generally into deeper water (figX). These shifts will place increasing pressure on a management system based around stable species distributions. The proportion of NEFMC managed benthivore species (righteye flounders, haddock) has declined over time in Mid-Atlantic waters according to bottom trawl surveys (figX), while the proportion of MAFMC managed planktivore (squids, mackerel, and butterfish) species has been increasing in New England waters in both surveys and landings (figsXX). Similarly, butterfish have been observed in Gulf of Maine common tern fledgling diets 2009-2011 and again in 2018 (see New England report). The downward trend in recreational species diversity in the Mid-Atlantic (figX) contrasts with an increase in recreational species diversity over time in New England (see New England report), although it is unclear to what extent this result is due to the aggregation of SAFMC species in the indicator itself. Nearshore NEAMAP survey indices show some similar patterns to offshore surveys, although with different magnitudes, perhaps reflecting seasonal importance of nearshore habitats; these patterns will be explored in more detail in future analyses. As temperature and ocean circulation indicators trend toward extremes, fishery management based on static stock areas will likely face continued changes in species distribution.

Observed changes at the base of the food web, including timing and community composition, affect productivity of protected and managed species in ways we do not yet fully understand. There is a trend of increasing primary production in the Mid-Atlantic, but this trend is primarily driven by increased summer production, which is due to warmer temperatures and increased bacterial remineralization and nutrient recycling (figX). This increased productivity is most likely from smaller-celled species that contribute less to fish production compared to larger phytoplankton. Current zooplankton trends in the Mid-Atlantic suggest a shift in timing towards a spring peak in the dominant species (figX), as well as a shift towards smaller-bodied copepods (figX). This suggests a possible return to conditions last observed during the 1990’s, when small bodied copepods dominated the zooplankton community, regime shifts in groundfish recruitment were observed (Peretti et al 2017), and north Atlantic right whales experienced lower birth rates (Greene et al. 2013). The timing of shifts in fish condition may be similar (figX), though potential mechanisms connecting adult fish condition to zooplankton patterns are unclear.

# Report Structure

The take home messages above synthesize the major messages of the report and refers to key figures. The figure format is explained in Fig \ref{fig:figformat}. The information in this report is organized around general ecosystem-level management objectives (Table \ref{tab:management-objectives}), and indicators related to these objectives are grouped into four general categories in the four sections below: economic and social, protected species, fish and invertebrates, and habitat quality and ecosystem productivity. Fish and invertebrates are aggregated into similar feeding categories (Table \ref{tab:species-groupings}) to evaluate ecosystem level trends in predators and prey. The spatial scale of indicators is either coastwide, Mid-Atlantic states (New York, New Jersey, Delaware, Maryland, Virginia, North Carolina), or at the Mid-Atlantic Bight (MAB) Ecosystem Production Unit (EPU, Fig. \ref{fig:EPU-MAP}) level. Each section begins with a summary of main messages with links to other sections, including any new information added at the request of the Council, and includes figures with brief descriptions of all current indicators. Technical methods documentation for the indicators is available [online](https://noaa-edab.github.io/tech-memo), and indicator data is also available [online](https://www.integratedecosystemassessment.noaa.gov/regions/northeast/Indicator-data#MidAtlanticIndicators).

```{r figformat, fig.cap = "Key to figures. Trend lines are shown when slope is significantly different from 0 at the p < 0.05 level. An orange line signifies an overall positive trend, and purple signifies a negative trend. To minimize bias introduced by small sample size, no trend is fit when N < 30. Dashed lines represent mean values of time series unless the indicator is an anomaly, in which case the dashed line is equal to 0. Shaded regions indicate the past ten years. If there are no new data for 2018, the shaded region will still cover this time period.", fig.height = 2.5, fig.width = 4}

m <- 0.1
x <- 1985:2018
y <-  m*x + rnorm(30, sd = 0.35)

data <- data.frame(x = x,
                  y = y)

#Define constants for figure plot
x.shade.max <- max(x)
x.shade.min <- x.shade.max - 9 
hline = mean(y)

#Plot series with trend 
ggplot2::ggplot(data = data,aes(x = x, y = y)) +
  #Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  geom_point(size = pcex) +
  scale_color_manual(aesthetics = "color")+
  guides(color = FALSE) +
  geom_hline(aes(yintercept = hline),
              size = hline.size,
             alpha = hline.alpha,
             linetype = hline.lty)+
  geom_line() +
  geom_gls() +
  scale_y_continuous(labels = function(l){trans = l / 1000})+
  scale_x_continuous(breaks = seq(1985, 2015, by = 5), expand = c(0.01, 0.01)) +
  ylab(expression("Invented Index, 10"^3*"widgets")) +
  xlab("Year") +
  theme_ts()

```

```{r management-objectives}

mng_obj <- data.frame("Objective Categories" = c("Seafood Production",
                                                 "Profits","Recreation",
                                                 "Stability","Social & Cultural",
                                                 "Biomass","Productivity",
                                                 "Trophic structure","Habitat"),
"Indicators reported here" = c("Landings by feeding guild","Revenue by feeding guild",
                               "Number of anglers and trips; recreational catch",
                               "Diversity indices (fishery and species)",
                               "Commercial and recreational reliance; social vulnerability",
                               "Biomass or abundance by feeding guild from surveys",
                               "Condition and recruitment of MAFMC managed species",
                               "Relative biomass of feeding guilds, primary productivity",
                               "Estimated habitat occurrence and potential wind-energy interactions"))

knitr::kable(mng_obj,
      col.names = c("Objective Categories","Indicators reported here"),
      caption = "Established ecosystem-scale objectives in the Mid-Atlantic Bight",
      #align = 'c',
      booktabs = T) %>%
 # kable_styling(latex_options = "striped") %>%
 # column_spec(c(2), width = c("25em")) %>%
  row_spec(0, bold = TRUE)
```

```{r species-groupings, warning=F}

# new table with all species listed by management entity
df <- ecodata::species_groupings %>%
  dplyr::select(SOE_18, COMNAME, Fed_Managed) %>%
  filter(SOE_18 != "Other") %>%
  distinct() %>%
  group_by(SOE_18, Fed_Managed) %>%
  summarize_all(funs(paste(na.omit(.), collapse = ", "))) %>%
  spread(Fed_Managed, COMNAME) %>%
  arrange(factor(SOE_18, levels = c("Apex Predator", "Piscivore", "Planktivore", "Benthivore", "Benthos")))
df<-df[-6,c(1,4,3,5,6)] %>%
  mutate_all(tolower)


knitr::kable(df, booktabs = TRUE, caption = 'Feeding guilds and management bodies.', 
             col.names = c("Guild", "MAFMC", "Joint", "NEFMC", "State or Other")) %>%
  kable_styling(font_size=10, latex_options=c("repeat_header", "scale_down", "hold_position")) %>%
  column_spec(1, width="2cm") %>%
  column_spec(2, width="5cm") %>%
  column_spec(3, width="2cm") %>%
  column_spec(4:5, width="5cm") %>%
  #column_spec(3, width="7.5cm") #%>%
  collapse_rows(columns = 1, latex_hline = "major", valign = "middle")

```

```{r EPU-MAP, fig.cap = "The Northeast Large Marine Ecosystem.", fig.width = 3.4, fig.asp = 1.25}

#Specify data frame with lat/lon locations for labels
epu_labels <- data.frame(EPU = c("Mid-Atlantic\n Bight",
                                 "Gulf of Maine",
                                 "Georges Bank"),
                         latitude = c(40,42.85,41),
                         longitude = c(-72.7,-69,-68.5))

#Map of NE LME
ggplot() +
  geom_sf(data = coast, size = map.lwd) +
  geom_sf(data = epu_sf, fill = "transparent", size = map.lwd) +
  coord_sf(crs = crs, xlim = xlims, ylim = ylims) +
  geom_text(data = epu_labels, aes(x = longitude,
                                    y = latitude,
                                    label = EPU),
            size = 3.2)+
  theme_map() +
  xlab("Longitude") +
  ylab("Latitude")

```

\newpage

# Economic and Social

The objectives of U.S. federal fishery management include providing benefits to the Nation in terms of seafood production and recreational opportunities, while considering economic efficiency and effects on coastal communities. The indicators in this section consider these objectives for commercial and recreational fishing sectors separately where possible. 

Despite mostly meeting fishery management objectives at the single species level (fig.X), long term declines in total seafood production and commercial revenue remain apparent (figX). Indicators highlight a declining diversity of recreational opportunities (fishing modes and species, figX). Further, coastal communities with high fishery engagement and reliance are dependent on a smaller number of species than historically (pX), these species are predominantly high valued shellfish vulnerable to increased ocean temperature and acidification.  

## Commercial sector (MAB)
Seafood landings overall and for MAFMC managed species declining (Fig. \ref{fig:total-landings}). Seafood landings for feeding guilds are also stable or declining (Fig. \ref{fig:comm-landings}), with the exception of a long term increase in MAFMC managed benthivores (scup, black sea bass, and tilefish). In 2018, landings of MAFMC managed planktivores increased to account for nearly all planktivore landings in the MAB, likely due to increased squid and decreased Atlantic herring landings.

```{r total-landings, fig.cap = paste0("Total commercial seafood landings (black) shown with ",region," managed seafood landings (red)."), fig.width = 5, fig.asp = 0.45}
#Get data for plotting

#Managed landings
managed_landings <- ecodata::comdat  %>%
  filter(str_detect(Var, paste0(council_abbr," managed species - Landings weight|JOINT managed species - Landings weight")),
         !str_detect(Var, "Other"),
         Time >= 1986,
         EPU == epu_abbr)

#Total landings
total_landings <- ecodata::comdat  %>%
  filter(!str_detect(Var, "managed species"),
         !str_detect(Var, "Other"),
         str_detect(Var, "Landings"),
         Time >= 1986,
         EPU == epu_abbr)

total_landings_agg <- total_landings %>%
  group_by(Time) %>%
  dplyr::summarise(Value = sum(Value)) %>% 
  mutate(Var = "Total",hline = mean(Value))
managed_landings_agg <- managed_landings %>%
  group_by(Time) %>%
  dplyr::summarise(Value = sum(Value)) %>% 
  mutate(Var = "Managed",hline = mean(Value))

landings_agg <- rbind(total_landings_agg, managed_landings_agg)

#Define constants for figure plot
x.shade.max <- 2018
x.shade.min <- 2009
series.col <- c("indianred","black")


ggplot(data = landings_agg)+
  
  #Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  geom_gls(aes(x = Time, y = Value,
               group = Var),
             alpha = trend.alpha, size = trend.size) +
  geom_line(aes(x = Time, y = Value, color = Var), size = lwd) +
  geom_point(aes(x = Time, y = Value, color = Var), size = pcex) +

  scale_y_continuous(labels = function(l){trans = l / 1000})+
  scale_x_continuous(breaks = seq(1985, 2015, by = 5), expand = c(0.01, 0.01)) +
  scale_color_manual(values = series.col, aesthetics = "color")+
  guides(color = FALSE) +
  ylab(expression("Landings (10"^3*"metric tons)")) +

  geom_hline(aes(yintercept = hline,
               
               color = Var),
           size = hline.size,
           alpha = hline.alpha,
           linetype = hline.lty) +
    theme_ts()

```

```{r comm-landings, fig.height=8, warning=F, fig.width = 5, fig.cap = paste0(council_abbr," managed species landings (red) and total commercial landings (black) by feeding guild.")}

#Assign feeding guild column for plotting with ggplot
landings <- rbind(managed_landings, total_landings) %>%
  mutate(feeding.guild = str_extract(Var,paste(feeding.guilds, collapse = "|")),
         grouping = factor(ifelse(str_detect(Var,council_abbr), "managed",
                                  ifelse(str_detect(Var, "JOINT"), "joint","total"))),
         Var = paste(word(feeding.guild), grouping)) %>% 
  mutate(feeding.guild = factor(feeding.guild, levels = feeding.guilds))

#Add JOINT landings to MANAGED landings and remove
landings[landings$Var ==  "Piscivore managed",]$Value <- landings[landings$Var ==  "Piscivore managed",]$Value + landings[landings$Var ==  "Piscivore joint",]$Value
landings <- landings %>%
  filter(Var != "Piscivore joint")  %>% 
  group_by(Var) %>% 
  mutate(hline = mean(Value))
  
#Define constants for figure plot
x.shade.max <- 2018
x.shade.min <- 2009
series.col <- c("indianred","black")


#facet names for titles
facet_names <- list("Apex predators" = expression("Apex predators"),
                    "Piscivores" = expression("Piscivores"),
                    "Planktivores" = expression("Planktivores"),
                    "Benthivores" = expression("Benthivores"),
                    "Benthos" = expression("Benthos"))

ggplot(data = landings,aes(x = Time, y = Value, color = grouping)) +
  
  #Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  
  #Test for trend and add lines
  geom_gls(aes(x = Time, y = Value,
               group = Var)) +
  
  #Add time series
  geom_line(size = lwd) +
  geom_point(size = pcex) +
  scale_color_manual(values = series.col, aesthetics = "color")+
  guides(color = FALSE) +
  geom_hline(aes(yintercept = hline,
                 color = grouping,
                 size = grouping),
             size = hline.size,
             alpha = hline.alpha,
             linetype = hline.lty)+

  #Facet 
  facet_wrap(feeding.guild~.,scales = "free_y", labeller = label, ncol = 1) +
  
  #Axis and theme
  scale_y_continuous(labels = function(l){trans = l / 1000})+
  scale_x_continuous(breaks = seq(1985, 2015, by = 5), expand = c(0.01, 0.01)) +
  ylab(expression("Landings, 10"^3*"metric tons")) +
  theme_facet() +
  theme(strip.text=element_text(hjust=0))


```


Revenue for MAFMC managed species has also declined over the long term (Fig. \ref{fig:comm-revenue}), with recent decreases in total revenue driven by decreased landings volume in the MAB which are outweighing increased prices for benthos, planktivores, and other species groups (Fig. \ref{fig:bennet}).

```{r comm-revenue, fig.width = 5, fig.asp = 0.45, fig.cap = "Total revenue for the region (black) and revenue from MAFMC managed species (red)."}

#Filtering and aggregation step
rev_agg <- ecodata::comdat %>% 
  filter(str_detect(Var, "Revenue"),
         !str_detect(Var, "prop|Other|NEFMC"), #Remove proportions, "Other" category species, NEFMC managed species in MAB
         EPU == epu_abbr,
         Time >= 1986) %>% 
  mutate(Status = ifelse(str_detect(Var, "Revenue weight"), 
                         "Managed","Total")) %>% #Create groups for aggregation
  group_by(Status, Time) %>% 
  dplyr::summarise(Total = sum(Value)) %>% 
  group_by(Status) %>% 
  mutate(hline = mean(Total))

series.col <- c("indianred","black")

#Plotting
ggplot(data = rev_agg) +
  
  #Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf)+  
  
  #lines
  geom_gls(aes(x = Time, y = Total,
               group = Status),
             alpha = trend.alpha, size = trend.size) +
  geom_line(aes(x = Time, y = Total, color = Status), size = lwd) +
  geom_point(aes(x = Time, y = Total, color = Status), size = pcex) +

  #axes
  scale_y_continuous(labels = function(l){trans = l / 1000000})+
  scale_x_continuous(breaks = seq(1985, 2015, by = 5), expand = c(0.01, 0.01)) +
      scale_color_manual(values = series.col, aesthetics = "color")+
  guides(color = FALSE) +
  ggtitle("Total revenue") +
  ylab(expression("Revenue (10"^6*"USD)")) +
  geom_hline(aes(yintercept = hline,
               color = Status),
           size = hline.size,
           alpha = hline.alpha,
           linetype = hline.lty) +
  theme_ts()

# rev_inline <- rev_agg %>% 
#   dplyr::select(Status, Time, Total) %>% 
#   spread(.,Status, Total) %>% 
#   mutate(percent_managed = round(Managed/Total * 100)) %>% 
#   filter(Time > latest_landings_data - 5, Time <= latest_landings_data) %>% 
#   pull(percent_managed)
# 
# min_rev_perc <- min(rev_inline)
# max_rev_perc <- max(rev_inline)
```


```{r bennet, fig.cap = paste0("Revenue change from the long-term mean in 2015 dollars (black), Price (PI), and Volume Indicators (VI) for commercial landings in the ",region,"."), fig.width = 6, fig.asp = 0.45}

#Filter data into two dataframes for plotting
indicators <- ecodata::bennet %>% 
  filter(EPU == epu_abbr,
         Var %in% c("VI EPU aggregate",
                    "PI EPU aggregate")) %>% 
  mutate(Var, Var = plyr::mapvalues(Var, from = c("VI EPU aggregate","PI EPU aggregate"),
                                    to = c("Volume","Price")))

revchange <- ecodata::bennet %>% 
  filter(EPU == epu_abbr,
         Var %in% c("REVCHANGE EPU aggregate"))

#custom bar fill color (color-blind friendly)
ind_fill <- c("#a6cee3", "#b2df8a")

#limits
y.lim <- c(-450,450)

#plot
ggplot(data = indicators)+
  
  #Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf)+
  
  geom_bar(aes(x=Time, y= Value, fill = Var), stat="identity")+
  scale_fill_manual(name = "Indicators", values = ind_fill) +
  geom_line(data = revchange, aes(x = Time, y = Value, colour="$"))+
  scale_colour_grey(name ="Revenue Change") +
  ggtitle("Bennet Indicator")+
  labs(y="Value $1,000,000 ($2015)") +
  scale_x_continuous(breaks = seq(1985, 2015, by = 5), expand = c(0.01, 0.01)) +
  scale_y_continuous(breaks = seq(y.lim[1], y.lim[2], by = 100), limits = y.lim, expand = c(0.01, 0.01)) +
  theme_ts() +
  theme(title = element_text(size = 10))

```

MAB fishing communities are heavily reliant on shellfish species (especially scallops) for the majority of revenue. This reliance likely represents heightened risk to fishing communities, particularly along the coasts of New York, New Jersey, and Virginia in ports which are highly engaged and reliant on commercial fishing. This risk is heightened by the high climate vulnerability of shellfish to ocean acidification (Fig. \ref{fig:MAB-comm-reliance-maps-outlined}).

```{r MAB-comm-reliance-maps-outlined, fig.height = 8, fig.cap = "Commercial engagement (total pounds landed, value landed, commercial permits and commercial dealers in a community) and reliance (per capita engagement) based on 2016 landings and the ACS running average of 2012-2016 census data."}
xmin = -78
xmax = -70
ymin = 35
ymax = 42
xlims <- c(xmin, xmax)
ylims <- c(ymin, ymax)

biv_col <- c("4" = "#ca0020", "3" = "#f4a582", "2" = "grey", "1" = "#92c5de", "0" = "#0571b0")

eng_rel <- ecodata::eng_rel %>%
    filter(PRIMARY_LATITUDE < 42) %>% 
  dplyr::rename(comeng = ComEng_NE16_ct,
                comrel = ComRel_NE16_ct,
                receng = RecEng_NE16_ct,
                recrel = RecRel_NE16_ct,
                Latitude = PRIMARY_LATITUDE,
                Longitude = PRIMARY_LONGITUDE,
                State = STATEABBR) %>% 
  mutate(comsum = sum(as.numeric(comeng), as.numeric(comrel))) %>% 
  mutate(comeng = factor(comeng, ordered = TRUE, levels = 1:4),
         comrel = factor(comrel, ordered = TRUE, levels = 1:4),
         receng = factor(receng, ordered = TRUE, levels = 1:4),
         recrel = factor(recrel, ordered = TRUE, levels = 1:4))



(com <- eng_rel %>% 
  dplyr::select(comeng, comrel, comsum, Latitude, Longitude) %>% 
  filter(comeng != 0 &
         comrel != 0) %>% 
ggplot() +
  geom_sf(data = coast, size = map.lwd) +
  geom_sf(data  = ne_states, size = map.lwd) +
  coord_sf(crs = crs, xlim = xlims, ylim = ylims) +
  geom_point(aes(x = Longitude, y = Latitude,
                 fill = comeng, size = comrel),
             color = "black",pch = 21) +
  guides(fill = guide_legend(override.aes = list(size = 5),
                            reverse = T,
                            title = "Engagement"),
         size = guide_legend(reverse = T,
                            title = "Reliance")) +
  scale_fill_manual(values = biv_col) +
  theme_map() +
  xlab("Longitude") +
  ylab("Latitude") +
      theme(legend.position = c(0.8, 0.25), 
        legend.box = "horizontal",
        legend.direction = "vertical",
        legend.key=element_blank(),
        legend.key.width = unit(0, "cm"))+
  ggtitle("Commercial Reliance & Engagement"))

# com + rec + plot_layout(ncol = 2) &
#   theme(plot.margin = unit(c(0, 0, 0, 0), "cm"))

```

Aquaculture production in the Mid-Atlantic is dominated by shellfish as well. Virginia continues to lead oyster production in the US, though numbers sold stabilized in 2017 after a long term increase. Reported concerns for Virginia growers within the past two years include an increase in wild spat fall preventing farmed oysters from entering higher-value markets. Maryland and New Jersey oyster production continues to increase (Fig. \ref{fig:oyster-aqua}).

```{r oyster-aqua, fig.width = 5, fig.asp = 0.45, fig.cap = "Oyster aquaculture production in terms of number of oysters sold from Virginia, Maryland, and New Jersey."}

aqua <- ecodata::aquaculture %>% 
  group_by(Var) %>% 
  mutate(hline = mean(Value)) %>%
  ungroup() %>% 
  mutate(Var = plyr::mapvalues(Var, from = c("md oyster harvest","nj oyster harvest","va oyster harvest"),
                                                    to  = c("MD","NJ","VA"))) %>%
  dplyr::rename(State = Var)

aqua$State <- factor(aqua$State, levels = c("VA","MD","NJ"))


ggplot() + 
  geom_segment(aes(x=2005,xend=2017,y=mean(aquaculture[aquaculture$Var == "va oyster harvest",]$Value),
                   yend=mean(aquaculture[aquaculture$Var == "va oyster harvest",]$Value)),
               size = hline.size,
           alpha = hline.alpha,
           linetype = hline.lty,
           color = "#1b9e77",
           inherit.aes = F) +
  geom_segment(aes(x=2012,xend=2016,y=mean(aquaculture[aquaculture$Var == "nj oyster harvest",]$Value),
                   yend=mean(aquaculture[aquaculture$Var == "nj oyster harvest",]$Value)),
               size = hline.size,
           alpha = hline.alpha,
           linetype = hline.lty,
           color = "#d95f02",
           inherit.aes = F) +
  geom_segment(aes(x=2012,xend=2017,y=mean(aquaculture[aquaculture$Var == "md oyster harvest",]$Value),
                   yend=mean(aquaculture[aquaculture$Var == "md oyster harvest",]$Value)),
               size = hline.size,
           alpha = hline.alpha,
           linetype = hline.lty,
           color = "#7570b3",
           inherit.aes = F) +
 #Highlight last ten years
  geom_line(data = aqua, aes(x = Time, y = Value, color = State), size = lwd) +
  geom_point(data = aqua,aes(x = Time, y = Value, color = State), size = pcex) +
  scale_color_manual(values = c(VA = "#1b9e77", MD = "#7570b3",NJ = "#d95f02")) +
  scale_x_continuous(breaks = seq(2005,2018,3),expand = c(0.01, 0.01)) +
  scale_y_continuous(labels = function(l){trans = l / 1000000})+
  ggtitle("Oyster harvest")+
  ylab(expression("Oysters sold (10"^6*" n)")) +
  xlab("")+
  theme_ts()

```

Commercial fleet diversity indices were updated with 2017 data. Current diversity remains near the long term average. 

## Recreational sector (Mid-Atlantic states)
All recreational indicators have been updated with new Marine Recreational Information Program (MRIP) data, and new indicators for recreational diversity are presented in this report at the request of the MAFMC. In contrast to the commercial seafood production trends, recreational seafood production shows an increasing trend since the mid-1990s with the updated MRIP data (Fig. \ref{fig:rec-landings}))

```{r rec-landings, fig.cap = paste0("Total recreational seafood harvest in the ",region," region."), fig.width = 5, fig.asp = 0.45}

landings_rec <- ecodata::recdat %>% 
  filter(EPU == region_abbr,
         Var == "Recreational Seafood") %>% 
  mutate(hline = mean(Value))

series.col <- "black"

ggplot(data = landings_rec)+
  
  #Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  geom_gls(aes(x = Time, y = Value,
               group = Var),
             alpha = trend.alpha, size = trend.size) +
  geom_line(aes(x = Time, y = Value, color = Var), size = lwd) +
  geom_point(aes(x = Time, y = Value, color = Var), size = pcex) +

  scale_y_continuous(labels = function(l){trans = l / 1000000})+
  scale_x_continuous(breaks = seq(1985, 2015, by = 5), expand = c(0.01, 0.01)) +
  scale_color_manual(values = series.col, aesthetics = "color")+
  guides(color = FALSE) +
  ggtitle("Recreational seafood harvest") +
  ylab(expression("Fish caught (10"^6*"n)")) +

  geom_hline(aes(yintercept = hline,
               
               color = Var),
           size = hline.size,
           alpha = hline.alpha,
           linetype = hline.lty) +
  theme_ts()

```


Updated indicators for recreational opportunities (effort days and number of anglers) show general increases since the 1990s, peaking in the late 2000s and declining since then. This is similar to previously reported trends (Fig. \ref{fig:rec-op}).

```{r rec-op, fig.width = 5, fig.asp = 1, fig.cap = paste0("Recreational effort, and number of recreational anglers in the ",region,".")}
recdat <- ecodata::recdat %>% 
  filter(EPU == region_abbr) %>% 
  group_by(Var) %>% 
  mutate(hline = mean(Value))

ylim_re <- c(2e7, 7e7)
ylim_rd <- c(1.75,2.75)
ylim_ra  <- c(1e6, 3.5e6)

# #Create dataframe for label locations
# label_loc <- data.frame(xloc = min(recdat$Time)+0.3,
#                         yloc = c(ylim_re[2]*0.975,
#                                  ylim_rd[2]*0.975,
#                                  ylim_ra[2]*0.975),
#                         labels = LETTERS[1:3],
#                         Var = c("Recreational Effort",
#                                 "Recreational fleet effort diversity across modes",
#                                 "Recreational anglers"))

series.col <- "black"
# x.shade.min <- max(recdat$Time, na.rm = T) - 9
# x.shade.max <- max(recdat$Time, na.rm = T)

rec_effort <- recdat %>% 
  filter(Var == "Recreational Effort") %>% 
  ggplot() + 
 #Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  #label
  # annotate("text", 
  #          x = label_loc[label_loc$Var == "Recreational Effort",]$xloc,
  #          y = label_loc[label_loc$Var == "Recreational Effort",]$yloc,
  #          label = label_loc[label_loc$Var == "Recreational Effort",]$labels,
  #          size = letter_size)+
  geom_gls(aes(x = Time, y = Value,
               group = Var),
             alpha = trend.alpha, size = trend.size) +
  geom_line(aes(x = Time, y = Value, color = Var), size = lwd) +
  geom_point(aes(x = Time, y = Value, color = Var), size = pcex) +
  
  scale_x_continuous(expand = c(0.01, 0.01)) +
  scale_y_continuous(labels = function(l){trans = l / 1000000}, limits = ylim_re)+
  scale_color_manual(values = series.col, aesthetics = "color")+
  guides(color = FALSE) +
  ggtitle("Recreational effort")+
  ylab(expression("Days fished (10"^6*" N)")) +
  xlab("")+
  geom_hline(aes(yintercept = hline,
               color = Var),
           size = hline.size,
           alpha = hline.alpha,
           linetype = hline.lty) +
  theme_ts() 



rec_anglers <- recdat %>% 
  filter(Var == "Recreational anglers") %>% 
  ggplot() + 
 #Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
    # annotate("text", 
    #        x = label_loc[label_loc$Var == "Recreational anglers",]$xloc,
    #        y = label_loc[label_loc$Var == "Recreational anglers",]$yloc,
    #        label = label_loc[label_loc$Var == "Recreational anglers",]$labels,
    #        size = letter_size)+
  geom_gls(aes(x = Time, y = Value,
               group = Var),
             alpha = trend.alpha, size = trend.size) +
  geom_line(aes(x = Time, y = Value, color = Var), size = lwd) +
  geom_point(aes(x = Time, y = Value, color = Var), size = pcex) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  scale_y_continuous(labels = function(l){trans = l / 1000000}, limits = ylim_ra)+
  scale_color_manual(values = series.col, aesthetics = "color")+
  guides(color = FALSE) +
  ggtitle("Recreational anglers") +
  ylab(expression("Anglers (10"^6*" n)")) +
  xlab("Time")+
  geom_hline(aes(yintercept = hline,
               color = Var),
           size = hline.size,
           alpha = hline.alpha,
           linetype = hline.lty) +
  theme_ts()

cowplot::plot_grid(rec_effort, 
                   #rec_div, 
                   rec_anglers, 
                   #rec_div_catch,
                   ncol = 1, 
                   align = "hv") +
    theme(plot.margin = unit(c(0.1, 0, 0, 0), "cm"))

```

Newly developed indicators for the diversity of recreational effort (i.e. access to recreational opportunities) by mode (party/charter boats, private boats, shore-based), and diversity of catch (NEFMC and MAFMC managed species only) show significant long-term downward trends. The downward effort diversity trend is driven by party/charter contraction (from a high of 24% of angler trips to 6% currently), with a shift towards shorebased angling. Effort in private boats remained stable between 36-37% of angler trips across the entire series. The long-term decrease in catch diversity in the Mid-Atlantic states contrasts with an increase in recreational catch diversity in New England states over the same time period; this trend requires further investigation as SAFMC managed species are not currently tracked separately (Fig. \ref{fig:rec-div})

```{r rec-div, fig.width = 5, fig.asp = 1, fig.cap = paste0("Recreational effort diversity, and diversity of recreational catch in the ",region,".")}

rec_div <- recdat %>% 
  filter(Var == "Recreational fleet effort diversity across modes") %>% 
  ggplot() + 
 #Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  # annotate("text", 
  #          x = label_loc[label_loc$Var == "Recreational fleet effort diversity across modes",]$xloc,
  #          y = label_loc[label_loc$Var == "Recreational fleet effort diversity across modes",]$yloc,
  #          label = label_loc[label_loc$Var == "Recreational fleet effort diversity across modes",]$labels,
  #          size = letter_size)+
  geom_gls(aes(x = Time, y = Value,
               group = Var),
             alpha = trend.alpha, size = trend.size) +
  geom_line(aes(x = Time, y = Value, color = Var), size = lwd) +
  geom_point(aes(x = Time, y = Value, color = Var), size = pcex) +
  ylim(ylim_rd)+
  scale_x_continuous(expand = c(0.01, 0.01)) +
  scale_color_manual(values = series.col, aesthetics = "color")+
  guides(color = FALSE) +
  ggtitle("Rec. fleet effort diversity")+
  ylab(expression("Effective Shannon")) +
  xlab("")+
  geom_hline(aes(yintercept = hline,
               color = Var),
           size = hline.size,
           alpha = hline.alpha,
           linetype = hline.lty) +
  theme_ts()

rec_div_catch <- recdat %>% 
  filter(Var == "Recreational Diversity of Catch") %>% 
  ggplot() + 
 #Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
    # annotate("text", 
    #        x = label_loc[label_loc$Var == "Recreational anglers",]$xloc,
    #        y = label_loc[label_loc$Var == "Recreational anglers",]$yloc,
    #        label = label_loc[label_loc$Var == "Recreational anglers",]$labels,
    #        size = letter_size)+
  geom_gls(aes(x = Time, y = Value,
               group = Var),
             alpha = trend.alpha, size = trend.size) +
  geom_line(aes(x = Time, y = Value, color = Var), size = lwd) +
  geom_point(aes(x = Time, y = Value, color = Var), size = pcex) +

  scale_x_continuous(expand = c(0.01, 0.01)) +
  scale_color_manual(values = series.col, aesthetics = "color")+
  guides(color = FALSE) +
  ggtitle("Rec. diversity of catch")+
  ylab(expression("Effective Shannon")) +
  xlab("Time")+
  geom_hline(aes(yintercept = hline,
               color = Var),
           size = hline.size,
           alpha = hline.alpha,
           linetype = hline.lty) +
  theme_ts()


cowplot::plot_grid(#rec_effort, 
                   rec_div, 
                   #rec_anglers, 
                   rec_div_catch,
                   ncol = 1, 
                   align = "hv") +
    theme(plot.margin = unit(c(0.1, 0, 0, 0), "cm"))

```


Recreational fishing is important to many Mid-Atlantic communities, particularly on Long Island, the  New Jersey shoreline and the Delaware Bay. Communities that are most socially and economically dependent (both engaged and reliant) on recreational fishing may face additional risk due this downward shift in diversity of recreational opportunity and species catch (Fig. \ref{fig:MAB-rec-engage}).

```{r MAB-rec-engage, fig.height = 8, fig.cap = "Recreational engagement (shore, private vessel and for-hire recreational fishing in a community) and reliance (per capita engagement) based on 2016 landings and the ACS running average of 2012-2016 census data."}
(rec <- 
  eng_rel %>% 
  dplyr::select(receng, recrel, Latitude, Longitude) %>% 
  filter(receng != 0,
         recrel != 0) %>% 
ggplot() +
  geom_sf(data = coast, size = map.lwd) +
  geom_sf(data  = ne_states, size = map.lwd) +
  coord_sf(crs = crs, xlim = xlims, ylim = ylims) +
  geom_point(aes(x = Longitude, y = Latitude,
                 fill = receng, size = recrel),
             color = "black",pch = 21) +
  guides(fill = guide_legend(override.aes = list(size = 5),
                            reverse = T,
                            title = "Engagement"),
         size = guide_legend(reverse = T,
                            title = "Reliance")) +
  scale_fill_manual(values = biv_col) +
  theme_map() +
  xlab("Longitude") +
  ylab("Latitude") +
  theme(legend.position = c(0.8, 0.25), 
        legend.box = "horizontal",
        legend.direction = "vertical",
        legend.key=element_blank(),
        legend.key.width = unit(0, "cm"))+
  ggtitle("Recreational Reliance & Engagement"))
```

\newpage

# Protected Species

Protected species include marine mammals (under the Marine Mammal Protection Act), Endangered and Threatened species (under the Endangered Species Act), and migratory birds (under the Migratory Bird Treaty Act). In the Northeast US, endangered/threatened species include Atlantic salmon, Atlantic and shortnose sturgeon, all sea turtle species, and 5 baleen whales. Fishery management objectives for protected species generally focus on reducing interactions between resource and protected species; here we report on the status of these interactions as well as indicating the potential for future interactions driven by observed and predicted ecosystem changes in the Northeast US region. This year four Unusual Mortality Events (UMEs) have been declared for three large whale species and two seal species, with several mortalities attributed to human interactions. Also, a marine mammal climate vulnerability assessment is currently underway and for Atlantic and Gulf of Mexico and will be reported on in future versions of this report.

## Sea Turtles (coastwide)
Sea turtles are known to be susceptible to climate and ecosystem changes, and their distribution is influenced by water temperature. Sea turtle diets contain a considerable amount of gelatinous zooplankton, which are also influenced by changes in the ecosystem. At present, management measures to reduce sea turtle-fishery interactions are limited to the regions with historical observations of sea turtles and based on historical ocean temperature distributions. However, changes in climate may cause turtles to shift northward into areas with heavy fishing, possibly resulting in increased bycatch, and necessitating new management measures.

## Whales (coastwide)
North Atlantic right whales are among the most endangered large whale populations in the world. Changes in right whale trends can have implications for fisheries management where fisheries interact with these whales. Additional management restrictions could have a large impact on fishing times, gears, etc. Although the population increased steadily from 1990 to 2011, it has decreased recently (FIg. \ref{fig:NARW-abundance}). From Pace et al 2017: “The probability that the population’s trajectory post-2010 was a decline was estimated at 99.99%.” Reduced survival rates of adult females and diverging abundance trends between sexes have also been observed. It is estimated that there are only about 100 adult females remaining in the population. Further, right whale distribution has changed since 2010. The reasons for these changes is unclear, but changes in climate and primary prey (*Calanus finmarchicus*) are suspected.
 
Three large whale Unusual Mortality Events (UMEs) have been declared for north Atlantic right whales, humpback whales, and minke whales. In all three cases human interaction appears to have contributed to increased mortalities, although investigations are not complete. Twenty right whale mortalities have been documented in 2017 and 2018 so far. Among the 20 right whale deaths observed in 2017 and 2018 thus far, 5 were due to vessel strike (1 in US waters, 4 in Canadian waters), 6 from entanglement (2 in Canadian gear, 1 in unknown gear, and 3 others found in US waters), and the rest from unknown causes. UMEs have also been declared for humpback whales during 2016-2018 and 2017-2018 minke whales due to elevated strandings. Necropsy investigations on 50% of humpback whales showed evidence of human interaction, either ship strike or entanglement. Similarly, 60% of minke whales show evidence of human interactions or infectious disease. In both cases the investigations are still underway and more research is needed to determine the causes of these UME.

```{r NARW-abundance, fig.width = 6, fig.asp = 0.55,fig.cap = "Right whale abundance estimates with 95\\% credible intervals. These values represent the estimated number of animals alive sometime during the year referenced and NOT at the end of the year referenced. Seventeen known deaths were recorded in 2017 (likely all from anthropogenic causes), but these deaths were not reflected in the 2017 estimate because those animals were alive sometime during the year."}
hline <- mean(narw[narw$Var == "right whale abundance median",]$Value)

ecodata::narw %>% 
  spread(.,Var,Value) %>% 
  ggplot() +
#Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  geom_line(aes(x = Time, y = `right whale abundance median`), size = lwd-0.75) +
  geom_point(aes(x = Time, y = `right whale abundance median`), size = pcex-0.75) +
  geom_errorbar(aes(x = Time,
                    ymin = `right whale abundance lcl`,
                  ymax = `right whale abundance ucl`), 
                width = 0.25) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  scale_color_manual(values = series.col, aesthetics = "color")+
  guides(color = FALSE) +
  ggtitle("NARW abundance") +
  ylab(expression("Abundance (n)")) +
  xlab("Time")+
  geom_hline(aes(yintercept = hline),
           color = "black",
           size = hline.size,
           alpha = hline.alpha,
           linetype = hline.lty) +
  theme_ts()
```

## Grey Seals (coastwide)
The number of grey seals (*Halichoerus grypus*) in U.S. waters has risen dramatically in the last 2 decades, with few  observed in the early 1990s to roughly 24,000 observed in southeastern Massachusetts in 2015 (Pace et al. in press). Roughly 30,000 - 40,000 gray seals were estimated in southeastern Massachusetts in 2015, using correction factors applied to seal counts visible in Google Earth imagery. As of 2016, the size of the grey seal population in Canada, which is part of the same stock as the grey seals in the U.S., was estimated to be roughly 425,000, and increasing by 4% a year. Pups born on Muskeget Island MA, currently the largest pupping site for gray seals in the U.S., were first observed in 1988 and now number over 3,500. Trends in pup production at U.S. colonies appear to be increasing, and it is likely that U.S. pup production is being supplemented each year by animals from Canada. A UME for both gray and harbor seals was declared in 2018, triggering an investigation into the cause of this event. Tests so far suggest phocine distemper virus as a potential cause, although the investigation is not yet complete.
 
Fisheries interactions have increased over the past 2 decades, with fewer than 10 total estimated grey seal interactions in 1993, to more than 1000 in 2013 and 2015, and with preliminary 2017 estimates over 900. Analysis of seal diet is currently underway using a variety of techniques (analysis of stomach contents, fatty acids, and DNA) to assess the potential impact of seal population growth on the ecosystem and important commercial fish species.

## Harbor porpoise (coastwide)
Harbor porpoise bycatch has resulted in fisheries closures in the past, but current bycatch levels suggest that management measures have been effective, reducing this fishery interaction. The 5-year mean bycatch has been below the maximum permitted level (Potential Biological Removal, PBR) since 2011 (Fig. \ref{fig:harbor-porpoise-bars}), and the 2016 and draft 2017 annual bycatch estimates are among the lowest in the time series. Recent compliance with the harbor porpoise take reduction plan and reduced fishing effort are thought to contribute to low bycatch estimates. Potential recent shifts in porpoise distribution could also be contributing to low bycatch and this will be explored during the coming year. A new draft of harbor porpoise abundance estimate suggests stable or increasing abundance of the Gulf of Maine/Bay of Fundy harbor porpoise stock. Recent analyses have examined regional harbor porpoise diet, and suggest that harbor porpoise are not typically feeding on fish caught in gillnets. However, the impact of ecosystem changes on bycatch, population, or distribution remain unclear.

```{r harbor-porpoise-bars, fig.width = 6, fig.asp = 0.55, fig.cap = "Harbor porpoise bycatch estimate (black) shown with Potential Biological Removal (red). Error bars indicate 95\\% confidence interval."}
ecodata::harborporpoise %>% 
  spread(.,Var,Value) %>% 
ggplot() +
    annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  geom_line(aes(x = Time, y = `harbor porpoise bycatch estimate`, color = "Harbor Porpoise Bycatch"), size = lwd) +
  geom_point(aes(x = Time, y = `harbor porpoise bycatch estimate`, color = "Harbor Porpoise Bycatch"), size = pcex) +
  geom_errorbar(aes(x = Time,
                    ymin = `harbor porpoise bycatch lo95ci`,
                  ymax = `harbor porpoise bycatch up95ci`,
                  color = "Harbor Porpoise Bycatch"), 
                width = 0.25)+
  geom_line(aes(x = Time, y = `harbor porpoise bycatch pbr`, color = "PBR"), size = lwd-0.1) +
  scale_color_manual("", values = c("Harbor Porpoise Bycatch" = "black", "PBR" = "red")) +
  guides(color = guide_legend(override.aes = list(shape = c(19,NA)))) +
  scale_x_continuous(expand = c(0.01, 0.01)) +
  ggtitle("Harbor porpoise bycatch") +
  ylab("Bycatch Estimate (n)") +
  theme_ts() +
  theme(legend.position = c(0.4, 0.85), legend.background = element_rect(
    fill = "transparent"), legend.text = element_text(size = 10))

```


# Fish and Invertebrates


# Habitat Quality and Ecosystem Productivity


# Contributors

