---
title: "State of the Ecosystem 2019: Mid-Atlantic"
output: pdf_document
---

```{r setup, include=FALSE}

#Default Rmd options
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = 'center') #allows for inserting R code into captions

#Plotting and data libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(ecodata)
library(here)
library(kableExtra)
library(ggrepel)
library(stringr)
library(patchwork)
library(grid)
library(ggiraph)
library(vegan)
library(rpart)

#GIS libraries
library(sf)
library(rgdal)
library(raster)
library(rnaturalearth)

#Data directories
image.dir <- here("images")
gis.dir <- here("gis")

#GIS directory
#gis.dir <- here::here("inst","extdata","gridded")

#General inline text input for report
#Council
council <- "Mid-Atlantic Fishery Management Council"
council_abbr <- "MAFMC"

#Region identifiers
epu <- "Mid-Atlantic Bight"
epu_abbr <- "MAB"
region <- "Mid-Atlantic"
region_abbr <- "MA" #Some commercial data organized by "MA" or "NE" regions, not by EPU 


```


```{r GIS-setup}

#CRS
crs <- "+proj=longlat +lat_1=35 +lat_2=45 +lat_0=40 +lon_0=-77 +x_0=0 +y_0=0 +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"

#Coastline shapefile
coast <- ne_countries(scale = 10,
                          continent = "North America",
                          returnclass = "sf") %>%
             sf::st_transform(crs = crs)

#State polygons
ne_states <- ne_states(country = "united states of america",
                                      returnclass = "sf") %>%
  sf::st_transform(crs = crs)

#high-res polygon of Maine
#new_england <- read_sf(gis.dir,"new_england")

#EPU shapefile
epu_sf <- ecodata::epu_sf %>% 
  filter(EPU %in% c("MAB","GB","GOM"))

#Map line parameters
map.lwd <- 0.4

# Set lat/lon window for maps
xmin = -77
xmax = -65
ymin = 36
ymax = 45
xlims <- c(xmin, xmax)
ylims <- c(ymin, ymax)

#Time series constants
shade.alpha <- 0.3
shade.fill <- "lightgrey"
lwd <- 1
pcex <- 2
trend.alpha <- 0.5
trend.size <- 2
hline.size <- 1
hline.alpha <- 0.35
hline.lty <- "dashed"
label.size <- 5
hjust.label <- 1.5
letter_size <- 4
feeding.guilds <- c("Apex Predator","Piscivore","Planktivore","Benthivore","Benthos")
x.shade.min <- 2009
x.shade.max <- 2018
#Function for custom ggplot facet labels
label <- function(variable,value){
  return(facet_names[value])
}
```

# Take Home Messages

The purpose of this report is to synthesize available information relevant to fishery management in the Mid-Atlantic portion of the US Northeast Shelf. This 2019 report highlights where management interventions have proven successful to achieve ecological objectives, but also the considerable challenges for management posed by climate change and increasing trade-offs across conservation, fishing, and other human activities in this region. Finally, we describe combinations of ecological signals that present opportunities for further integrated research and possibly creative management solutions. 

Some management interventions for ecosystems and fisheries have proven beneficial in the Mid-Atlantic and throughout the Northeast US Shelf. For example, management limiting nutrient inputs has significantly improved water quality in Chesapeake Bay (fig.X). This trend, which has been driven by improvements to dissolved oxygen content and water clarity, has positive implications for aquaculture development and for managed species using the Bay as nursery habitat.  Also encouraging is a continued decrease in harbor porpoise bycatch, which for this year is the lowest on record in the region, and an apparent increase in the harbor porpoise population (p. X).

However, there are multiple signals indicating challenges to meeting management goals. Despite mostly meeting fishery management objectives at the single species level (fig.X), long term declines in total seafood production and commercial revenue remain apparent (figX). Indicators highlight a declining diversity of recreational opportunities (fishing modes and species, figX). Further, coastal communities with high fishery engagement and reliance are dependent on a smaller number of species than historically (pX), these species are predominantly high valued shellfish vulnerable to increased ocean temperature and acidification. 

Fisheries are significantly affecting protected species, particularly large whale abundance. Changes in fishery management to address this issue could have considerable consequences for the fixed gear fishing sector, particularly for pot gear such as for lobster. The recovery of north Atlantic right whales is jeopardized by anthropogenic mortality (figX). Unusual Mortality Events (UMEs) have been declared for north Atlantic right whales, humpback whales, and minke whales, and preliminary investigations suggest fishing gear entanglement has played a primary role in these mortalities (pX). Gray seal bycatch mortality has also risen in recent years, with an estimated annual mortality that has often totaled more than 1000 animals and a UME has been declared for both gray and harbor seals that may be due to phocine distemper virus (pX). Shifts in ecosystem conditions may be contributing to increased interaction between right whales and fishing gear and could be playing a role in other protected species trends.

Several climate and ecosystem observations are trending towards unprecedented levels.  Sea surface and bottom temperatures across the Northeast US shelf are still among the fastest warming waters globally (figX) with implications for species physiology, productivity, distribution, and community composition. Globally, 2018 was the 4th hottest year on record and the last four years were the warmest on record. In the Mid-Atlantic, 2018 summer sea surface temperatures were the third highest on record, however temperatures during the other seasons were near or slightly below average.  Bottom temperature measurements also continue to have a significant long term increasing trend (figX). Ocean circulation is changing as well. The position of the northern edge of the Gulf Stream has been moving to the north since the late 1950’s, with an increasing rate since 2009. The most northerly positions ever recorded were observed over the most recent years (2014-2017, figX). A more northerly Gulf Stream position is associated with warmer ocean temperature in the Northeast US shelf (Zhang and Vallis, 2007) and increased sea surface height along the U.S. east coast (Goddard et al. 2015). 

The management implications of these ocean changes vary by region and species, and are not fully understood at this point. Changes in the distribution of managed fish species continue, with aggregate trends on the entire Northeast Shelf towards the northeast and generally into deeper water (figX). These shifts will place increasing pressure on a management system based around stable species distributions. The proportion of NEFMC managed benthivore species (righteye flounders, haddock) has declined over time in Mid-Atlantic waters according to bottom trawl surveys (figX), while the proportion of MAFMC managed planktivore (squids, mackerel, and butterfish) species has been increasing in New England waters in both surveys and landings (figsXX). Similarly, butterfish have been observed in Gulf of Maine common tern fledgling diets 2009-2011 and again in 2018 (see New England report). The downward trend in recreational species diversity in the Mid-Atlantic (figX) contrasts with an increase in recreational species diversity over time in New England (see New England report), although it is unclear to what extent this result is due to the aggregation of SAFMC species in the indicator itself. Nearshore NEAMAP survey indices show some similar patterns to offshore surveys, although with different magnitudes, perhaps reflecting seasonal importance of nearshore habitats; these patterns will be explored in more detail in future analyses. As temperature and ocean circulation indicators trend toward extremes, fishery management based on static stock areas will likely face continued changes in species distribution.

Observed changes at the base of the food web, including timing and community composition, affect productivity of protected and managed species in ways we do not yet fully understand. There is a trend of increasing primary production in the Mid-Atlantic, but this trend is primarily driven by increased summer production, which is due to warmer temperatures and increased bacterial remineralization and nutrient recycling (figX). This increased productivity is most likely from smaller-celled species that contribute less to fish production compared to larger phytoplankton. Current zooplankton trends in the Mid-Atlantic suggest a shift in timing towards a spring peak in the dominant species (figX), as well as a shift towards smaller-bodied copepods (figX). This suggests a possible return to conditions last observed during the 1990’s, when small bodied copepods dominated the zooplankton community, regime shifts in groundfish recruitment were observed (Peretti et al 2017), and north Atlantic right whales experienced lower birth rates (Greene et al. 2013). The timing of shifts in fish condition may be similar (figX), though potential mechanisms connecting adult fish condition to zooplankton patterns are unclear.

# Report Structure

The take home messages above synthesize the major messages of the report and refers to key figures. The figure format is explained in Fig \ref{fig:figformat}. The information in this report is organized around general ecosystem-level management objectives (Table \ref{tab:management-objectives}), and indicators related to these objectives are grouped into four general categories in the four sections below: economic and social, protected species, fish and invertebrates, and habitat quality and ecosystem productivity. Fish and invertebrates are aggregated into similar feeding categories (Table \ref{tab:species-groupings}) to evaluate ecosystem level trends in predators and prey. The spatial scale of indicators is either coastwide, Mid-Atlantic states (New York, New Jersey, Delaware, Maryland, Virginia, North Carolina), or at the Mid-Atlantic Bight (MAB) Ecosystem Production Unit (EPU, Fig. \ref{fig:EPU-MAP}) level. Each section begins with a summary of main messages with links to other sections, including any new information added at the request of the Council, and includes figures with brief descriptions of all current indicators. Technical methods documentation for the indicators is available [online](https://noaa-edab.github.io/tech-memo), and indicator data is also available [online](https://www.integratedecosystemassessment.noaa.gov/regions/northeast/Indicator-data#MidAtlanticIndicators).

```{r figformat, fig.cap = "Key to figures. Trend lines are shown when slope is significantly different from 0 at the p < 0.05 level. An orange line signifies an overall positive trend, and purple signifies a negative trend. To minimize bias introduced by small sample size, no trend is fit when N < 30. Dashed lines represent mean values of time series unless the indicator is an anomaly, in which case the dashed line is equal to 0. Shaded regions indicate the past ten years. If there are no new data for 2018, the shaded region will still cover this time period."}

m <- 0.1
x <- 1985:2018
y <-  m*x + rnorm(30, sd = 0.35)

data <- data.frame(x = x,
                  y = y)

#Define constants for figure plot
x.shade.max <- max(x)
x.shade.min <- x.shade.max - 9 
hline = mean(y)

#Plot series with trend 
ggplot2::ggplot(data = data,aes(x = x, y = y)) +
  #Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  geom_point(size = pcex) +
  scale_color_manual(aesthetics = "color")+
  guides(color = FALSE) +
  geom_hline(aes(yintercept = hline),
              size = hline.size,
             alpha = hline.alpha,
             linetype = hline.lty)+
  geom_line() +
  geom_gls() +
  theme_classic() + 
  scale_y_continuous(labels = function(l){trans = l / 1000})+
  scale_x_continuous(breaks = seq(1985, 2015, by = 5), expand = c(0.01, 0.01)) +
  ylab(expression("Invented Index, 10"^3*"widgets")) +
  xlab("Year")



```

```{r management-objectives}

mng_obj <- data.frame("Objective Categories" = c("Seafood Production",
                                                 "Profits","Recreation",
                                                 "Stability","Social & Cultural",
                                                 "Biomass","Productivity",
                                                 "Trophic structure","Habitat"),
"Indicators reported here" = c("Landings by feeding guild","Revenue by feeding guild",
                               "Number of anglers and trips; recreational catch",
                               "Diversity indices (fishery and species)",
                               "Commercial and recreational reliance; social vulnerability",
                               "Biomass or abundance by feeding guild from surveys",
                               "Condition and recruitment of MAFMC managed species",
                               "Relative biomass of feeding guilds, primary productivity",
                               "Estimated habitat occurrence and potential wind-energy interactions"))

knitr::kable(mng_obj,
      col.names = c("Objective Categories","Indicators reported here"),
      caption = "Established ecosystem-scale objectives in the Mid-Atlantic Bight",
      #align = 'c',
      booktabs = T) %>%
 # kable_styling(latex_options = "striped") %>%
 # column_spec(c(2), width = c("25em")) %>%
  row_spec(0, bold = TRUE)
```

```{r species-groupings, warning=F}

# new table with all species listed by management entity
df <- ecodata::species_groupings %>%
  dplyr::select(SOE_18, COMNAME, Fed_Managed) %>%
  filter(SOE_18 != "Other") %>%
  distinct() %>%
  group_by(SOE_18, Fed_Managed) %>%
  summarize_all(funs(paste(na.omit(.), collapse = ", "))) %>%
  spread(Fed_Managed, COMNAME) %>%
  arrange(factor(SOE_18, levels = c("Apex Predator", "Piscivore", "Planktivore", "Benthivore", "Benthos")))
df<-df[-6,c(1,4,3,5,6)] %>%
  mutate_all(tolower)


knitr::kable(df, booktabs = TRUE, caption = 'Feeding guilds and management bodies.', 
             col.names = c("Guild", "MAFMC", "Joint", "NEFMC", "State or Other")) %>%
  kable_styling(font_size=10, latex_options=c("repeat_header", "scale_down", "hold_position")) %>%
  column_spec(1, width="2cm") %>%
  column_spec(2, width="5cm") %>%
  column_spec(3, width="2cm") %>%
  column_spec(4:5, width="5cm") %>%
  #column_spec(3, width="7.5cm") #%>%
  collapse_rows(columns = 1, latex_hline = "major", valign = "middle")

```

```{r EPU-MAP, fig.cap = "The Northeast Large Marine Ecosystem.", fig.width = 3.4, fig.asp = 1.25}

#Specify data frame with lat/lon locations for labels
epu_labels <- data.frame(EPU = c("Mid-Atlantic\n Bight",
                                 "Gulf of Maine",
                                 "Georges Bank"),
                         latitude = c(40,42.85,41),
                         longitude = c(-72.7,-69,-68.5))

#Map of NE LME
ggplot() +
  geom_sf(data = coast, size = map.lwd) +
  geom_sf(data = epu_sf, fill = "transparent", size = map.lwd) +
  coord_sf(crs = crs, xlim = xlims, ylim = ylims) +
  geom_text(data = epu_labels, aes(x = longitude,
                                    y = latitude,
                                    label = EPU),
            size = 3.2)+
  theme_map() +
  xlab("Longitude") +
  ylab("Latitude")

```

# Economic and Social

The objectives of U.S. federal fishery management include providing benefits to the Nation in terms of seafood production and recreational opportunities, while considering economic efficiency and effects on coastal communities. The indicators in this section consider these objectives for commercial and recreational fishing sectors separately where possible. 

Despite mostly meeting fishery management objectives at the single species level (fig.X), long term declines in total seafood production and commercial revenue remain apparent (figX). Indicators highlight a declining diversity of recreational opportunities (fishing modes and species, figX). Further, coastal communities with high fishery engagement and reliance are dependent on a smaller number of species than historically (pX), these species are predominantly high valued shellfish vulnerable to increased ocean temperature and acidification.  

## Commercial sector (MAB)
Seafood landings overall and for MAFMC managed species are stable or declining, with the exception of a long term increase in MAFMC managed benthivores (scup, black sea bass, and tilefish). In 2018, landings of MAFMC managed planktivores increased to account for nearly all planktivore landings in the MAB, likely due to increased squid and decreased Atlantic herring landings (commercial landings fig).

Revenue for MAFMC managed species has also declined over the long term, with recent decreases in total revenue driven by decreased landings volume in the MAB which are outweighing increased prices for benthos, planktivores, and other species groups (commercial revenue fig).

MAB fishing communities are heavily reliant on shellfish species (especially scallops) for the majority of revenue. This reliance likely represents heightened risk to fishing communities, particularly along the coasts of New York, New Jersey, and Virginia in ports which are highly engaged and reliant on commercial fishing. This risk is heightened by the high climate vulnerability of shellfish to ocean acidification (commercial engagement/reliance fig).

Aquaculture production in the Mid-Atlantic is dominated by shellfish as well. Virginia continues to lead oyster production in the US, though numbers sold stabilized in 2017 after a long term increase. Reported concerns for Virginia growers within the past two years include an increase in wild spat fall preventing farmed oysters from entering higher-value markets. Maryland and New Jersey oyster production continues to increase (aquaculture fig).

Commercial fleet diversity indices were updated with 2017 data. Current diversity remains near the long term average.  (no fig? fig?)

## Recreational sector (Mid-Atlantic states)
All recreational indicators have been updated with new Marine Recreational Information Program (MRIP) data, and new indicators for recreational diversity are presented in this report at the request of the MAFMC. In contrast to the commercial seafood production trends, recreational seafood production shows an increasing trend since the mid-1990s with the updated MRIP data. (rec landings fig)

Updated indicators for recreational opportunities (effort days and number of anglers) show general increases since the 1990s, peaking in the late 2000s and declining since then. This is similar to previously reported trends. (rec opportunities fig--effort and angler days)

Newly developed indicators for the diversity of recreational effort (i.e. access to recreational opportunities) by mode (party/charter boats, private boats, shore-based), and diversity of catch (NEFMC and MAFMC managed species only) show significant long-term downward trends. The downward effort diversity trend is driven by party/charter contraction (from a high of 24% of angler trips to 6% currently), with a shift towards shorebased angling. Effort in private boats remained stable between 36-37% of angler trips across the entire series. The long-term decrease in catch diversity in the Mid-Atlantic states contrasts with an increase in recreational catch diversity in New England states over the same time period; this trend requires further investigation as SAFMC managed species are not currently tracked separately. (rec diversity fig)

Recreational fishing is important to many Mid-Atlantic communities, particularly on Long Island, the  New Jersey shoreline and the Delaware Bay. Communities that are most socially and economically dependent (both engaged and reliant) on recreational fishing may face additional risk due this downward shift in diversity of recreational opportunity and species catch. (rec engagement reliance fig)

# Protected Species


# Fish and Invertebrates


# Habitat Quality and Ecosystem Productivity


# Contributors

