---
title: "null"
csl: plos.csl
fontsize: 10pt
output:
  pdf_document:
    includes:
      in_header: latex/header.tex
    keep_tex: yes
  html_document:
    df_print: paged
  word_document: default
link-citations: yes
geometry: left=2cm, right=2cm, top=2cm, bottom=3cm, footskip = .5cm
subparagraph: yes
bibliography: SOE2020.bib
---

```{r setup, include=FALSE}

# library(tint)
# # invalidate cache when the package version changes
# knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tint'))
# options(htmltools.dir.version = FALSE)

#Default Rmd options
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      dev = "cairo_pdf",
                      warning = FALSE,
                      fig.width = 4,
                      fig.asp = 0.45,
                      fig.align = 'center'
                      ) #allows for inserting R code into captions

#Plotting and data libraries
#remotes::install_github("noaa-edab/ecodata@0.1.0") #change to 2020 ecodata version for release
library(tidyverse)
library(tidyr)
library(ecodata)
library(here)
library(kableExtra)

```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-setup.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-GIS-setup.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-setup.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-setup.R")}
```

\setcounter{page}{4}

# Introduction

## About This Report

The purpose of this report is to synthesize ecosystem information to better meet fishery management objectives, and to update the MAFMC’s Ecosystem Approach to Fishery Management (EAFM) risk assessment. The major messages of the report are synthesized in the 2-page summary and synthesis themes are illustrated on page 3. The information in this report is organized into two sections; [performance measured against ecosystem-level management objectives](#performance-relative-to-fishery-management-objectives) (Table \ref{tab:management-objectives}), and potential [risks to meeting fishery management objectives](#risks-to-meeting-fishery-management-objectives) (Climate change and Other Ocean Uses).

## Report structure

The two main sections contain subsections for each management objective or potential risk. Within each subsection, we first review indicator trends, and the status of the most recent year relative to a threshold (if available) or relative to the long-term average. Second, we synthesize results of other indicators and information to outline potential implications for management (i.e., connecting indicator(s) status to management and why an indicator(s) is important) . For example, if there are multiple drivers related to an indicator trend, which drivers may be more or less supported by current information, and which, if any, can be affected by management action? Similarly, which risk indicators warrant continued monitoring to evaluate whether regime shifts or ecosystem reorganization are likely? We emphasize that these implications are intended to represent testable hypotheses at present, rather than “answers,” because the science behind these indicators and syntheses continue to develop.  

Detailed technical methods documentation^[https://NOAA-EDAB.github.io/tech-doc] and indicator data^[https://github.com/NOAA-EDAB/ecodata] are available online. The details of standard figure formatting (Fig. \ref{fig:docformat}a), categorization of fish and invertebrate species into feeding groups (Table \ref{tab:species-groupings}), and definitions of ecological production units (EPUs, including the Mid-Atlantic Bight, MAB; Fig. \ref{fig:docformat}b) are provided at the end of the document. 

```{r management-objectives}

mng_obj <- data.frame("Objective Categories" = c("Seafood Production",
                                                 "Profits","Recreation",
                                                 "Stability","Social & Cultural",
                                                 "Protected Species",
                                                 "Biomass","Productivity",
                                                 "Trophic structure","Habitat"),
"Indicators reported here" = c("Landings by feeding guild","Revenue decomposed to price and volume",
                               "Days fished; recreational catch",
                               "Diversity indices (fishery and species)",
                               "Community engagement/reliance status",
                               "",
                               "Biomass or abundance by feeding guild from surveys",
                               "Condition and recruitment of managed species, Primary productivity",
                               "Relative biomass of feeding guilds, Zooplankton",
                               "Estuarine and offshore habitat conditions"))

knitr::kable(mng_obj, linesep = "",
      col.names = c("Objective Categories","Indicators reported here"),
      caption = "Established ecosystem-scale objectives in the Mid-Atlantic Bight",
      #align = 'c',
      booktabs = T) %>%
  kable_styling(latex_options = "hold_position", "scale_down") %>%
 # column_spec(c(2), width = c("25em")) %>%
  row_spec(0, bold = TRUE) %>%
 # group_rows("Provisioning/Cultural", 1,4) %>%
 # group_rows("Supporting/Regulating", 5,9)
  pack_rows("Provisioning/Cultural Services", 1,6) %>%
  pack_rows("Supporting/Regulating Services", 7,10)


```
 
# Performance relative to fishery management objectives

In this section, we examine indicators related to broad, ecosystem-level fishery management objectives. We also provide hypotheses on the implications of these trends—*why* we are seeing them, what’s driving them, and potential or observed regime shifts or changes in ecosystem structure. Identifying multiple drivers, regime shifts, and potential changes to ecosystem structure, as well as identifying the most vulnerable resources, can help managers determine whether we can do anything differently to meet objectives and how to prioritize for upcoming issues/risks.

## Seafood Production 

### Indicator: Landings

Total commercial landings (black) and MAFMC’s managed landings (red) in the Mid-Atlantic Bight (MAB)  continue to trend downward. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-comdat-total-landings.R")}
```

The downward trend is most significant in the benthos (scallops and clams) group. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-comdat-commercial-landings.R"), fig.asp=1.5}
```

Total recreational landings (retained fish presumed to be eaten) are also down in the MAB. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-recdat-landings.R")}

```

Recreational shark landings show an increase in pelagic sharks over the past decade, with a sharp decrease in 2018 and 2019. This is likely influenced by regulatory changes implemented in 2018 intended to rebuild shortfin mako stocks. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-rec_hms.R")}

```

Aquaculture production is not yet included in total seafood landings, but we are working toward including it in future reports. Available aquaculture production of oysters for a subset of Mid-Atlantic states is trending upward.^[https://noaa-edab.github.io/catalog/aquaculture.html]

### Implications

Declining commercial and recreational landings can be driven by many interacting factors, including combinations of ecological and stock production, management actions, market conditions, and environmental change. While we cannot evaluate all possible drivers at present, here we evaluate the extent to which ecosystem overfishing (total landings exceeding ecosystem productive capacity), stock status, and system biomass trends may play a role. 

#### Ecosystem Overfishing Indices 
Thresholds for ecosystem-level overfishing based on system production characteristics have been proposed, and are applied here for the MAB.  Based on either the ratio of total catch to total primary production (Fogarty Index), or to total catch per unit area (Ryther Index), ecosystem overfishing in the MAB is unlikely based on the proposed thresholds, so ecosystem overfishing is unlikely to be a major factor driving decreased landings. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-fogarty.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-ryther.R")}
```

Fisheries catches are sustained by a lower proportion of the ecosystem's primary production now than in the past, particularly when compared with the 1970s when each of these indices suggest that ecosystem overfishing may have occurred.
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-ppr.R")}
```

#### Stock Status
Stock status is meeting objectives for most MAFMC managed stocks, so stock status and associated management constraints are unlikely to be driving decreased landings. To better address the role of management in future reports, we could examine how the total allowable catch (TAC) and the percentage of the TAC taken for each species has changed through time. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-stock-status.R"), fig.width = 7.5, fig.asp = 0.5}
```

#### System Biomass
Although aggregate biomass trends derived from scientific resource surveys are mostly stable in the MAB, spring piscivores and fall benthos show long-term increases. (The NEAMAP Fall 2020 survey was completed and is included here; NEFSC surveys were not completed in 2020). While managed species make up varying proportions of aggregate biomass, trends in landings are not mirroring shifts in the overall trophic structure of survey-sampled fish and invertebrates. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-aggregate-biomass.R"), fig.width=8, fig.asp = 0.75}
```

#### Effect on Seafood Production
Because ecosystem overfishing seems unlikely, stock status is mostly acceptable, and aggregate biomass trends appear stable, the decline in commercial landings is most likely driven by market dynamics affecting the landings of surfclams and ocean quahogs, as quotas are not binding for these species. 

Climate change also seems to be shifting the distribution of surfclams and ocean quahogs, resulting in areas with overlapping distributions and increased mixed landings . Given the regulations governing mixed landings, this could become problematic in the future and is currently being evaluated by the Council. 

The decline in recreational seafood landings stems from other drivers.  Some of the decline, such as that for recreational shark landings, is driven by management intended to reduce fishing mortality on mako sharks. However, NOAA Fisheries’ Marine Recreational Information Program survey methodology was updated in 2018, so it is unclear whether the record-low landings for 2018 are driven by changes in fishing behavior or the change in the survey methodology.  

Other environmental changes  require monitoring as they may become important drivers of landings in the future:
 
- Climate trending to uncharted territory (see [Climate Risks section](#climate-and-ecosystem-productivity))  
- Stocks are shifting distribution throughout the region  
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-species-dist.R"), fig.asp=1}
```  
- Some ecosystem composition and production changes (see [Stability section](#stability))  
- Engagement declining (see [Social and Cultural section](#social-and-cultural))  

## Commercial Profits 

### Indicators: revenue (a proxy for profits), with price and volume components  
Total commercial revenue (black) has increased over the long term, but the trend may be reversing, with recent total revenue below the long-term average. The MAFMC-managed species revenue (red) has continued its downward trend, with recent years near a time-series low. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-comdat-comm-revenue.R")}
```

Total revenue trends (black line) mirror those for the benthos (clams and scallops; orange in separate volume and price indicator plots) group over the past decade. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-bennet2.R"), fig.width = 7.5, fig.asp = 0.5}
```

### Implications 
Revenue earned by harvesting resources is a function of both the quantity landed of each species and the prices paid for landings. Beyond monitoring yearly changes in revenue, it is even more valuable to determine what drives these changes:  harvest levels, the mix of species landed, price changes, or a combination of these. The Bennet Indicator decomposes revenue change into two parts, one driven by changing quantities (volumes), and a second driven by changing prices. 

The Bennet indicator demonstrates that increasing total revenue early in the time series is due to increasing quantities landed, which offset declining prices. Recent declines in prices contributed to falling revenue as quantities landed did not increase enough to counteract declining prices. 

Changes in other indicators, particularly those related to climate change, require monitoring as they may become important drivers of revenue in the future:

-  Surfclams and ocean quahogs are  sensitive to warming ocean temperatures and ocean acidification.   
-  Acidification levels in surfclam summer habitat are approaching, but not yet at, levels affecting surfclam growth (see [Climate Risks section](#climate-and-ecosystem-productivity)).  

## Recreational Opportunities 

### Indicators: Days fished, fleet diversity  
Recreational effort (days fished) has no significant long term trend, with current days fished near the long-term average. However, recreational fleet diversity has declined over the long term. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-recdat-effort.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-recdat-diversity.R")}

```

### Implications
The absence of a long-term trend in recreational days fished suggests relative stability in the overall number of recreational opportunities in the MAB. However,  the decline in recreational fleet diversity suggests a potentially reduced range of opportunities. 

The downward effort diversity trend is driven by party/charter contraction (from a high of 24% of angler trips to 7% currently), and a shift toward shorebased angling. Effort in private boats remained stable between 36-37% of angler trips across the entire series.

Changing recreational fleet diversity can be considered to maintain recreational opportunities. Shore anglers will have access to different species and when the same species, typically smaller fish. Many states have developed shore based regulations where the minimum size is lower than in other areas and sectors to maintain opportunities in the shore angling sector. 


## Stability 

### Indicators: fishery fleet and catch diversity, ecological component diversity 
While there are many potential metrics of stability, we use diversity indices as a first check to evaluate overall stability in fisheries and ecosystems. In general, diversity that remains constant over time suggests a similar capacity to respond to change over time. A significant change in diversity over time does not necessarily indicate a problem or an improvement, but does indicate a need for further investigation. We examine commercial and recreational fleet and species catch diversity, and diversity in zooplankton, larval, and adult fish. 

#### Fishery Diversity 
Commercial fishery fleet count, fleet diversity, and permit diversity has been stable over time in the MAB, with current values near the long-term average. This indicates similar commercial fleet composition and species targeting opportunities over time. Commercial fisheries are relying on fewer species relative to the mid -90s, but consistent with, and near, the long term average. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-commercial-div.R"), fig.asp=1}
```

As noted [above](#recreational-opportunities)recreational fleet effort diversity is unstable (declining). However, recreational species catch diversity is stable and has been at or above the long term average in 7 of the last 10 years. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-recdat-div-catch.R")}
```

#### Ecological Diversity

Ecological diversity indices show mixed trends. Zooplankton diversity is increasing in the MAB. Larval fish and adult fish diversity indices are stable over time, with current values near the long-term average. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-zoo-diversity.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-ichthyo-diversity.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-exp-n.R"), fig.width=5}
```

 
### Implications
Fleet diversity indices are used by the MAFMC to evaluate stability objectives as well as risks to fishery resilience and maintaining equity in access to fishery resources [@gaichas_implementing_2018]. 

Stability in commercial fleet diversity metrics suggests stable capacity to respond to the current range of fishing opportunities. 

Declining recreational fleet effort diversity, as noted [above](#recreational-opportunities), indicates that the party/charter boat sector continues to contract, with shoreside angling becoming more important, as a percentage of recreational days fished.

Stability in recreational species catch diversity has been maintained by a different set of species over time. A recent increase in ASMFC and SAFMC managed species in recreational catch is helping to maintain diversity in the same range that MAFMC and NEFMC species supported in the 1990s.

Ecological diversity indices can provide insight into ecosystem structure. Changes in ecological diversity over time may indicate altered ecosystem structure with implications for fishery productivity and management.

Increasing zooplankton diversity is driven by the declining dominance of the calanoid copepod *Centropages typicus*, with a similar composition of other zooplankton species. 

Stable larval and adult fish diversity indicates the same overall number and evenness over time, but doesn’t rule out species substitutions (e.g., warm-water replacing cold-water). Stable but variable larval diversity can indicate interannual changes in a dominant species.

In the MAB, existing diversity indicators suggest overall stability in the fisheries and ecosystem components examined. However, declining recreational fleet diversity suggests a potential loss in the range of recreational fishing opportunities, and increasing zooplankton diversity is due to the declining dominance of an important species, suggesting change in the zooplankton community that warrants continued monitoring to determine if managed species are affected. 

## Social Vulnerability 

### Indicators: Social vulnerability in commercial and recreational fishing communities
Social vulnerability measures social factors that shape a community’s ability to adapt to change and does not consider gentrification pressure vulnerability. Communities that ranked medium-high or above for one or more of the following indicators: poverty, population composition, personal disruption, or labor force structure, are highlighted in  red. 

Commercial fishery engagement measures the number of permits, dealers, and landings in a community, while reliance expresses these numbers based on the level of fishing activity relative to the total population of a community.  
In 2020, we reported that the number of highly engaged Mid-Atlantic commercial fishing communities had declined over time, and engagement scores had also declined in medium-highly engaged communities. Here we focus on the top ten most engaged, and top ten most reliant commercial fishing communities and their associated  social vulnerability.  Barnegat Light and Cape May, NJ, and Reedville, VA are highly engaged and reliant with medium-high to high social vulnerability.   
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-commercial-engagement.R"), fig.width = 7.5, fig.asp = 0.75}
```

Recreational fishery engagement measures shore, private vessel, and for-hire fishing activity while reliance expresses these numbers based on fishing effort relative to the population of a community. Of the nine recreational communities that are most engaged and reliant, Avon, Ocracoke and Hatteras, NC and Barnegat Light and Cape May, NJ scored medium-high or above for social vulnerability. 

Both commercial and recreational fishing are important activities in  Montauk, NY; Barnegat Light, Cape May, and Point Pleasant Beach, NJ; and Ocracoke and Rodanthe, NC, meaning some of these communities may be impacted simultaneously by commercial and recreational regulatory changes. . Of these communities, three scored medium-high or above for social vulnerability.

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-recreational-engagement.R"), fig.width = 7.5, fig.asp = 0.75}
```


### Implications
These plots provide a snapshot of the relationship between social vulnerability and the most highly engaged and most highly reliant commercial and recreational fishing communities in the Mid-Atlantic. Similar plots are used to inform the annual California Current Ecosystem Status Report. These communities may be vulnerable to changes in fishing patterns due to regulations and/or climate change. When any of these communities are also experiencing social vulnerability, they may have lower ability to successfully respond to change. These indicators may also point to communities that are vulnerable to environmental justice issues. Additional analysis related to ecosystem shifts and National Standard 8 of the Magnuson-Stevens Act is ongoing. 


## Protected Species
Protected species include marine mammals protected under the Marine Mammal Protection Act, endangered and threatened species protected under the Endangered Species Act, and migratory birds protected under the Migratory Bird Treaty Act. In the Northeast U.S., endangered/threatened species include Atlantic salmon, Atlantic and shortnose sturgeon, all sea turtle species, and five baleen whales. Fishery management objectives for protected species generally focus on reducing threats and on habitat conservation/restoration; here we report on the status of these actions as well as indicating the potential for future interactions driven by observed and predicted ecosystem changes in the Northeast U.S. region.
Protected species objectives include managing bycatch to remain below potential biological removal (PBR) thresholds, recovering endangered populations, and monitoring unusual mortality events (UMEs).

### Indicators: bycatch, population (adult and juvenile) numbers, mortalities
BAverage indices for both harbor porpoise and gray seal bycatch are below current PBR thresholds, meeting management objectives.  
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-harborporpoise.R")}
```

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-grayseal.R")}
```

The North Atlantic right whale population was on a recovery trajectory until 2010, but has since declined. Reduced survival rates of adult females and diverging abundance trends between sexes have also been observed. It is estimated that there are only about 100 adult females remaining in the population. 
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-narw-abundance.R")}
```
North Atlantic right whale calf counts have also been declining. In 2018 there were zero observed new calves, and a drop in annual calves roughly mirrors the abundance decline , however seven new calves were born in 2019. Preliminary 2020 observations of 12 calves have been recorded as of January 2021.  
```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-NARW-calf-abundance.R")}
```

This year, four Unusual Mortality Events (UMEs) continued, three for large whales and one for gray and harbor seals.
The three large whale Unusual Mortality Events (UMEs) are ongoing for North Atlantic right whales, humpback whales, and minke whales. 

Since 2017, the total UME right whale mortalities includes 32 dead stranded whales, 11 in the US and 21 in Canada. When alive but seriously injured whales (14) are taken into account, 46 individual whales are included in the UME. During 2020 two mortalities were documented, however, recent research suggests that many mortalities go unobserved and the true number of mortalities are about three times the count of the observed mortalities. The primary cause of death is “human interaction” from entanglements or vessel strikes. 

Coastal bottlenose dolphin stocks off North Carolina and Virginia are listed as depleted, so a take reduction team met in 2019 and has been evaluating and implementing some of the team's consensus recommendations. 

Also, a UME for both gray and harbor seals was declared in 2018 due to a high number of mortalities thought to be caused by phocine distemper virus.

### Implications
Bycatch management measures have been implemented to maintain bycatch below thresholds. The downward trend in harbor porpoise bycatch can also be due to a decrease in  harbor porpoise abundance in US waters, which has reduced fisheries overlap; and a decrease in gillnet effort. The increasing trend in gray seal bycatch may be related to an increase in the gray seal population (U.S. pup counts).

The number of gray seals in U.S. waters has risen dramatically in the last three  decades. Based on a survey conducted in 2016, the size of the gray seal population in the U.S. during the breeding season was approximately 27,000 animals, while in Canada the population,was estimated to be roughly 425,000. A survey conducted in 2021 in both countries will provide updated estimates of abundance. The population in Canada is increasing at roughly 4% per year, and contributing to rates of increase in the U.S., where the number of pupping sites has increased from 1 in 1988 to 9 in 2019. Mean rates of increase in the number of pups born at various times since 1988 at four of the more data-rich pupping sites (Muskeget, Monomoy, Seal, and Green Islands) ranged from -0.2% (95\%CI: -2.3 - 1.9\%) to 26.3% (95\%CI: 21.6 - 31.4\%). These high rates of increase provide further support for the hypothesis that seals from Canada are continually supplementing the breeding population in U.S. waters.  

Strong evidence exists to suggest that interactions between right whales and the offshore lobster gear in the U.S. and snow crab gear in Canada is contributing substantially to the decline of the species. Further, right whale distribution has changed since 2010. The reasons for these changes is unclear, but changes in climate and the distribution of their primary prey (*Calanus finmarchicus*) are suspected causes.

The UMEs are under investigation and are likely the result of multiple drivers. For all three large whale UMEs, human interaction appears to have contributed to increased mortalities, although investigations are not complete. An investigation into the cause of the seal UME so far suggests phocine distemper virus as a potential cause. 

A marine mammal climate vulnerability assessment is currently underway for Atlantic and Gulf of Mexico populations and will be reported on in future versions of this report.


# Risks to meeting fishery management objectives

## Climate and Ecosystem Productivity

### Climate Change Indicators: ocean currents, temperature, heatwaves, acidification
Regional ocean current indicators remain at unprecedented levels. The Gulf Stream Index shows an all time northward progression of this current in 2019. 
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-gsi.R")}
```
In 2019, we also observed the second lowest proportion of Labrador Shelf water entering the Gulf of Maine since 1978. 
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_NE.Rmd-slopewater.R")}
```
Ocean temperatures continue to warm at both the surface and bottom. Warming is not seasonally uniform, however: spring 2020 was cooler than average on portions of the shelf.
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-bottom-temp.R")}
```

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-seasonal-sst-anom-gridded.R"), fig.width = 8, fig.height = 7}
```

Chesapeake Bay also experienced a warmer-than-average winter and a cooler-than-average spring in 2020, relative to the previous decade.  Water temperatures returned to average during the summer and slightly warmer than average October through December. 

The MAB experienced frequent ocean heatwaves of moderate intensity in 2020 that extended well into December, as observed in Chesapeake Bay.
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-heatwave-year.R"), fig.asp=.6}
```

New measurements show areas of low pH (7.8) in summer in Mid-Atlantic habitats occupied by surfclams and sea scallops. This seasonal pH minimum is associated with cold pool subsurface and bottom water, which is cut off from ventilation with surface water due to strong stratification. 

```{r, fig.asp=1}
knitr::include_url("https://github.com/NOAA-EDAB/ecodata/raw/master/docs/images/Seasonal%20pH%20on%20MAB%20shelf%20-%20Grace%20Saba.jpg")
```

Nearshore we see lower pH because of freshwater input - during the fall overturn the pH increases because of the input of the salty water, primary productivity increases pH, but increased respiration decreases pH

The MAB cold pool area was near average in 2018, with smallest areas associated with record-warm years (2012).

### Ecosystem Productivity Indicators: primary production, zooplankton, forage fish, fish condition
Increased temperatures drive increased rates of photosynthesis in summer
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-pp-monthly.R"), fig.width=8}
```

Increased overall primary production from summer production, larger winter blooms in 2020?
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-chl-weekly.R"), fig.width = 5, fig.asp = 0.9}
```

Climatology of seasonal size fractions confirms that summer production dominated by smaller size classes (pico and nano)
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-weekly-phyto-size.R"), fig.asp=.6}
```

Less efficient transfer to higher trophic levels

Trends in gelatinous zoops, krill are the same across EPUs as last year, data was updated to 2019. Long term increase almost all regions.
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-zoo-strat-abun.R"), fig.width=8, fig.asp=.25}
```

Larger zooplankton (Cal fin) had above average abundance in 2018-2019.
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-zoo-abund.R")}
```

Trends in Cal fin overwintering stage important as forage for right whales and forage fish.
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/LTL_MAB.Rmd-calanus-stage.R"), fig.width=8, fig.asp=.25}
```


Forage anomaly from zooplankton data, no apparent trend in MAB
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-forage-anomaly.R")}
```


Forage energy density updated; lower now for herring than in 80s-90s, other species seasonal difference but near previous estimates.
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-energy-density.R"), fig.width = 7.5, fig.asp = 0.5}
```


Temperature and zooplankton changes impact fish condition for different species, impacts to fisheries and markets are under investigation.
```{r, fig.width = 7.5, fig.asp = 0.5}
knitr::include_url("https://github.com/NOAA-EDAB/ecodata/raw/master/docs/images/condition_proj.png")
```


### Ecosystem Structure Indicators: distribution shifts, diversity, predators
As noted in the [Landings "Why" section above](#landings), stocks are shifting distribution throughout the region. In aggregate, fish stocks are moving northeast along the shelf and into deeper waters.  
Zooplankton diversity is increasing in the MAB, while larval fish and adult fish diversity indices are stable over time with current values near the long term average (see [Diversity Indicators section, above](#diversity)).  
New indicators for shark populations, combined with information on grey seals (see [Protected Species "Why" section, above](#protected-species)), suggests predator populations are stable to increasing in the MAB.

```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-observed-sharks.R")}
```


```{r}
#code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/macrofauna_MAB.Rmd-hms-cpue-sharks.R"),
#sp_cat<- read.csv(here::here("data-raw/hms-mrip/hms_sp_category.csv"))
sp_cat<- read.csv("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/data-raw/hms-mrip/hms_sp_category.csv")
ecodata::hms_cpue %>% 
  filter(str_detect(Var, "SHARK")) %>% 
  rename(COMMON_POP = Var) %>% 
  left_join(sp_cat) %>% 
  group_by(Year, SP_CATEGORY) %>% 
  summarise(Value = sum(Value)) %>% 
  rename(Var = SP_CATEGORY) %>% 
  filter(!Var == "NA") %>% 
  ggplot()+
 ggplot2::annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  ggplot2::geom_point(aes(x=Year, y = Value, color = Var))+
  ggplot2::geom_line(aes(x=Year, y = Value, color = Var))+
  #ggplot2::facet_wrap(~Var, scales = "free")+
  ggplot2::ggtitle("HMS POP SHARK CPUE")+
  ggplot2::ylab("Number per Haul")+
  ecodata::theme_ts()
```

### Habitat Climate Vulnerability
New habitat climate vulnerability analysis links black sea bass, scup, and summer flounder to several highly vulnerable nearshore habitats from salt marsh through shallow estuarine and marine reefs.

### Implications

Climate-managed species links

The warm winter may have affected key Chesapeake Bay fishery resources during a critical period. Results of the Maryland juvenile striped bass survey, conducted by the Maryland Department of Natural Resources (MDNR), showed low recruitment success in 2020, about fivefold below the long-term average. This low recruitment event may have been caused by a mismatch in striped bass (Morone saxatilis) larval and prey abundance due to the warm winter conditions, resulting in reduced larval survival. Warm winters typically trigger an early plankton bloom, including key copepod prey, which dies out before striped bass larvae are present in the tributary (Millette et al. 2019). 

Conversely, warmer winter temperatures may have reduced overwintering mortality of Chesapeake Bay blue crabs (Callinectes sapidus). Calculations done by MDNR based on data from the annual Bay-wide winter dredge survey indicate that blue crabs experienced the lowest overwintering mortality ever calculated (2020 Chesapeake Bay Blue Crab Advisory Report). Previous studies have demonstrated the correlation between winter water temperature and blue crab survival in the Chesapeake Bay (Bauer & Miller 2010, Hines et al. 2010, Rome et al. 2005). 

In addition to winter water temperature, survival of early life stages of striped bass in the Chesapeake Bay is strongly correlated with freshwater flow (Martino & Houde 2010, Millette et al. 2019, North & Houde 2003); high-flow regimes push zooplankton prey downstream, where they get trapped in the estuarine turbidity maximum with striped bass larvae. In low-flow years, zooplankton prey are less likely to match up with striped bass larvae in space and time, reducing larval survival and recruitment success. The combined effects of warm winter temperatures and low flow in 2020 may be the primary cause of the low recruitment success observed by the MDNR juvenile striped bass survey. 

Conversely, increased salinity often results in high juvenile oyster abundance in the Chesapeake Bay (Kimmel et al. 2014). In Maryland, the 2020 MDNR fall oyster survey documented above-average spatsets along the Eastern Shore as expected given the high salinity; however, the Western Shore did not fare as well, suggesting that local environmental conditions are also important.

Ocean acidification also has different implications, depending on the species and life stage. The decrease of pH decreases the feeding rates. Surfclam more sensitive than sea scallops. Lab studies have found a surfclam threshold metabolic depression in a pH range of 7.46-7.28. 
Ocean pH from one season of data (mostly 2019) hasn't reached the threshold observed for surfclams in lab studies so far--BUT we haven’t yet tested thresholds for other species or for multiple life stages, specifically larval stages that are typically more vulnerable to OA.

More variable system--not uniformly warmer, more variable too and this makes a wider range of conditions for biota to deal with

2012 still warmest on record, now papers published on impacts to give insight into what the implications of heatwaves can be
Lobster was a big one--timing of fishing, markets, etc
Cobia habitat shifts (and other species)--seen off RI now

Fish Condition Links to Environmental Drivers

General Additive Models (GAMs) were used to test how measures of fishing pressure, stock abundance and individual environmental variables performed in explaining the changes of fish condition (fatness) over time. Some species such as Acadian redfish, butterfish and winter flounder were more affected by fishing pressure and stock size, whereas other species such as weakfish, windowpane and American plaice may be more affected by local bottom temperatures and zooplankton. These relationships can potentially provide insights on which species may be more vulnerable to environmental changes such as climate change, as well as what biomass changes may be expected from certain species given current environmental conditions. Correlations were examined between environmental drivers, and as expected there were strong temperature correlations between seasons as well as correlations between temperature and zooplankton indices. Planned future work includes building full GAM models for each fish species, and linking fish condition to socio-economic models to assess whether fish condition impacts the value generated by that species.
 
Socio-Economic Models
 
Economic theory and empirical analyses have highlighted that many factors can affect the price of fish, including the total quantity of fish in the market (sometimes including internationally), increased demand around holidays, time the fish was in storage, and other issues that either affect the quality of the fish or the amount of fish available for purchase. We plan on empirically exploring whether fish condition is a quality metric people care about. Understanding whether people care about fish condition will help us more holistically understand the impacts of condition change on society, if any.

## Other Ocean Uses: Offshore Wind

### Indicators: revenue in lease areas, development timeline, survey overlap
Commercial fishery revenue in proposed offshore wind lease areas ranges from x-y% of total revenue for MAFMC managed species.
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-wind-revenue.R")}
```

Offshore wind projects may cover more than 1.5 million acres by 2029.
```{r, code=readLines("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/chunk-scripts/human_dimensions_MAB.Rmd-wind-dev-speed0.R")}
```

Scientific survey area overlap with offshore wind lease areas ranges from 2-25% across ecosystem, shellfish, fish, and protected species surveys.
```{r}
knitr::include_url("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/docs/images/SurveysWindFarmAreas_20201229.jpg")
```

### Implications
Fishery revenue could be displaced to other areas or foregone if all wind projects are built.  
Displaced fishing effort can alter fishing methods and change habitat, species (managed and protected), and fleet interactions.  
Right whales may be displaced and could face possible prey impacts via changed oceanography.  
Current plans for rapid buildout of offshore wind in a patchwork of areas spreads the impacts differentially throughout the region.  
```{r, fig.asp=1.5}
knitr::include_url("https://raw.githubusercontent.com/NOAA-EDAB/ecodata/master/docs/images/MidAtlantic_2021119-01.jpg")
```

Offshore wind will pose a risk to scientific data collection for surveys and increased uncertainty for management decision making across physical, low trophic, fish, and protected species.




# Contributors

**Editors** (NOAA NMFS Northeast Fisheries Science Center, NEFSC): Sarah Gaichas, Kimberly Bastille, Geret DePiper, Kimberly Hyde, Scott Large, Sean Lucey, Chris Orphanides 

**Contributors** (NEFSC unless otherwise noted): Andy Beet, Patricia Clay, Lisa Colburn, Geret DePiper, Michael Fogarty, Paula Fratantoni, Kevin Friedland, Sarah Gaichas, Avijit Gangopadhyay (School for Marine Science and Technology, University of Massachusetts Dartmouth), James Gartland (Virginia Institute of Marine Science), Glen Gawarkiewicz (Woods Hole Oceanographic Institution), Sean Hardison, Kimberly Hyde, John Kosik, Steve Kress and Don Lyons (National Audubon Society’s Seabird Restoration Program), Ruth Boettcher (Virginia Department of Game and Inland Fisheries), Young-Oh Kwon and Zhuomin Chen (Woods Hole Oceanographic Institute), Sean Lucey, Chris Melrose, Ryan Morse, Kimberly Murray, Chris Orphanides, Richard Pace, Charles Perretti, Vincent Saba, Laurel Smith, John Walden, Harvey Walsh, Mark Wuenschel.  


\newpage 

# Document Orientation

The figure format is illustrated in Fig \ref{fig:docformat}a. Trend lines are shown when slope is significantly different from 0 at the p < 0.05 level. An orange line signifies an overall positive trend, and purple signifies a negative trend. To minimize bias introduced by small sample size, no trend is fit for < 30 year time series. Dashed lines represent mean values of time series unless the indicator is an anomaly, in which case the dashed line is equal to 0. Shaded regions indicate the past ten years. If there are no new data for 2018, the shaded region will still cover this time period. The spatial scale of indicators is either coastwide, Mid-Atlantic states (New York, New Jersey, Delaware, Maryland, Virginia, North Carolina), or at the Mid-Atlantic Bight (MAB) Ecosystem Production Unit (EPU, Fig. \ref{fig:docformat}b) level.

```{r docformat, fig.cap = "Document orientation. a. Key to figures. b.The Northeast Large Marine Ecosystem.",  fig.width = 8, fig.height = 2.5}
#fig.subcap= c('Key to figures.', 'The Northeast Large Marine Ecosystem.'), out.width = '.49\\linewidth', fig.show="hold" 
# FIgure orientation subfigure
m <- 0.1
x <- 1985:2018
y <-  m*x + rnorm(30, sd = 0.35)

data <- data.frame(x = x,
                  y = y)

#Define constants for figure plot
x.shade.max <- max(x)
x.shade.min <- x.shade.max - 9 
hline = mean(y)

#Plot series with trend 
psample <- ggplot2::ggplot(data = data,aes(x = x, y = y)) +
  #Highlight last ten years
  annotate("rect", fill = shade.fill, alpha = shade.alpha,
      xmin = x.shade.min , xmax = x.shade.max,
      ymin = -Inf, ymax = Inf) +
  geom_point(size = pcex) +
  scale_color_manual(aesthetics = "color")+
  guides(color = FALSE) +
  geom_hline(aes(yintercept = hline),
              size = hline.size,
             alpha = hline.alpha,
             linetype = hline.lty)+
  geom_line() +
  geom_gls() +
  scale_y_continuous(labels = function(l){trans = l / 1000})+
  scale_x_continuous(breaks = seq(1985, 2015, by = 5), expand = c(0.01, 0.01)) +
  ylab(expression("Invented Index, 10"^3*"widgets")) +
  xlab("Time") +
  labs(tag = "a")  +
  theme_ts()


# EPU map subfigure
# Set lat/lon window for map
xmin = -78
xmax = -65
ymin = 36
ymax = 45
xlims <- c(xmin, xmax)
ylims <- c(ymin, ymax)

#CRS
crs <- "+proj=longlat +lat_1=35 +lat_2=45 +lat_0=40 +lon_0=-77 +x_0=0 +y_0=0 +datum=NAD83 +no_defs +ellps=GRS80 +towgs84=0,0,0"

#Specify data frame with lat/lon locations for labels
epu_labels <- data.frame(EPU = c("Mid-Atlantic\n Bight",
                                 "Gulf of Maine",
                                 "Georges Bank"),
                         latitude = c(40,42.85,41),
                         longitude = c(-72.7,-69,-68.5))

#Map of NE LME
epumap <- ggplot() +
  geom_sf(data = coast, size = map.lwd) +
  geom_sf(data = epu_sf, fill = "transparent", size = map.lwd) +
  coord_sf(crs = crs, xlim = xlims, ylim = ylims) +
  geom_text(data = epu_labels, aes(x = longitude,
                                    y = latitude,
                                    label = EPU),
            size = 1.7)+
  theme_map() +
  scale_x_continuous(breaks = seq(-78, -65, by = 4), expand = c(0.01, 0.01)) +
  xlab("Longitude") +
  ylab("Latitude") +
  labs(tag = "b")

psample + plot_spacer() + epumap + plot_layout(ncol = 3, widths=c(3,.5,4.5))

```

Fish and invertebrates are aggregated into similar feeding categories (Table \ref{tab:species-groupings}) to evaluate ecosystem level trends in predators and prey. 

```{r species-groupings, warning=F}

# new table with all species listed by management entity
df <- ecodata::species_groupings %>%
  dplyr::select(SOE.20, COMNAME, Fed.Managed) %>%
  filter(SOE.20 != "Other") %>%
  distinct() %>%
  group_by(SOE.20, Fed.Managed) %>%
  summarize_all(funs(paste(na.omit(.), collapse = ", "))) %>%
  spread(Fed.Managed, COMNAME) %>%
  arrange(factor(SOE.20, levels = c("Apex Predator", "Piscivore", "Planktivore", "Benthivore", "Benthos")))
df<-df[c(1,3,2,4,5)] %>%
  mutate_all(tolower)


knitr::kable(df, booktabs = TRUE, caption = 'Feeding guilds and management bodies.', 
             col.names = c("Guild", "MAFMC", "Joint", "NEFMC", "State or Other")) %>%
  kable_styling(font_size=10, latex_options=c("repeat_header", "scale_down", "hold_position")) %>%
  row_spec(0,bold=TRUE) %>%
  column_spec(1, width="2cm") %>%
  column_spec(2, width="4cm") %>%
  column_spec(3, width="2cm") %>%
  column_spec(4, width="5cm") %>%
  column_spec(5, width="6cm") %>%
  #column_spec(3, width="7.5cm") #%>%
  collapse_rows(columns = 1, latex_hline = "major", valign = "middle")

```


\newpage 

# References
